---
title: "01_Cleaning"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true 
output_dir: Output 
--- 
```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

source("../Scripts/functions.R")
```

This script will contain the code necessary to clean the data stored in the `/RawData` folder. The first data cleaning section will focus on mortality data, using data retried from CDC Wonder. Documentation for exact queries used to pull specific data are contained within each datafile itself. 

The second cleaning section will focus on population data from the California Department of Finance. 

Functions used to clean data are contained in the `/Scripts/functions.R` document. 


# All-Cause Mortality by Year 

```{r}
#H 
all_cause_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/All_Cause/H.xlsx",
                          col_names = TRUE)
all_cause_h <- H_ucd_clean(all_cause_h, "all_cause")  
glimpse(all_cause_h)

#NH 
all_cause_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/All_Cause/NH.xlsx",
                          col_names = TRUE) 

all_cause_nh <- NH_ucd_clean(all_cause_nh, "all_cause") 
glimpse(all_cause_nh)  

#Binding rows 
all_cleaned <- bind_rows(
  all_cause_h,
  all_cause_nh
)

glimpse(all_cleaned)
#Cleaning final merge

all_cleaned <- all_cleaned %>%
  filter(age_group != "Not Stated") %>% 
  mutate(
    sex = str_to_lower(sex),
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ), 
    age_group = str_trim(age_group),
    age_group = case_when(
      age_group %in% c("< 1 year", "1-4 years") ~ "0-4",
      age_group == "5-9 years"   ~ "5-9",
      age_group == "10-14 years" ~ "10-14",
      age_group == "15-19 years" ~ "15-19",
      age_group == "20-24 years" ~ "20-24",
      age_group == "25-29 years" ~ "25-29",
      age_group == "30-34 years" ~ "30-34",
      age_group == "35-39 years" ~ "35-39",
      age_group == "40-44 years" ~ "40-44",
      age_group == "45-49 years" ~ "45-49",
      age_group == "50-54 years" ~ "50-54",
      age_group == "55-59 years" ~ "55-59",
      age_group == "60-64 years" ~ "60-64",
      age_group == "65-69 years" ~ "65-69",
      age_group == "70-74 years" ~ "70-74",
      age_group == "75-79 years" ~ "75-79",
      age_group == "80-84 years" ~ "80-84",
      age_group == "85-89 years" ~ "85-89",
      age_group %in% c("90-94 years", "95-99 years", "100+ years") ~ "90+",
      TRUE ~ NA_character_
    )
  )
all_cleaned <- all_cleaned %>% 
  group_by(year, sex, age_group, cod, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = FALSE),
    .groups = "drop"
  ) %>% 
  arrange(year,age_group)

# checking NA values
all_cleaned %>% 
  filter(is.na(deaths))  

head(all_cleaned)

write.csv(all_cleaned, "../CleanData/all_cause_UCD.csv")

  
``` 

The only NA death values (they were suppressed due to data constraints) are among Black children, usually 5-9 years old. This makes some sense, as they are the smallest racial group in CA.  



# Homicide  

## By age
```{r}
# By Age Group 
hom_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Age/H.xlsx", col_names = TRUE)
hom_h <- ucd_age_clean_h(hom_h, "homicide")  
glimpse(hom_h) 

#we have 23 obs, which is equivalent to one hispanic row for each age group. looks good. 

#NH
hom_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Age/NH.xlsx", col_names = TRUE)
hom_nh <- ucd_age_clean_nh(hom_nh, "homicide")  
glimpse(hom_nh)

age_homicide <-  bind_rows(
  hom_nh,
  hom_h
) 


#Cleaning final merge

age_homicide <- race_age_format(age_homicide) 
glimpse(age_homicide)

#adding population data 
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarize(population = sum(population))
head(pop) 

age_hom2 <- age_homicide %>%
  left_join(pop, by = c("race_eth", "age_group"))
head(age_hom2)

write.csv(age_hom2, "../CleanData/homicide_UCD_age_groups.csv")
```

## By year: 
```{r}
#By Year (15-24 year olds)
hom_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Year/H.xlsx", col_names = TRUE)
hom_h <- ucd_year_clean_h(hom_h, "homicide")  
glimpse(hom_h) 

#NH
hom_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Year/NH.xlsx", col_names = TRUE)
hom_nh <- ucd_year_clean_nh (hom_nh, "homicide")  
glimpse(hom_nh)

year_homicide <-  bind_rows(
  hom_nh,
  hom_h
) 


#Cleaning final merge

year_homicide <- year_homicide %>% 
  mutate(
    race_eth = case_when(
      race == "Black or African American" ~ "black",
      race == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    )
  )
glimpse(year_homicide) 
unique(year_homicide$race_eth)

#adding population data 
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    age_group %in% c("15-19","20-24"),
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, year) %>% 
  summarize(population = sum(population))
head(pop) 

year_hom2 <- year_homicide %>%
  left_join(pop, by = c("race_eth", "year"))
head(year_hom2)

write.csv(year_hom2, "../CleanData/homicide_UCD_yearly.csv")

```

# MVC 

## By Age 

```{r}
mvc_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Age/NH.xlsx", col_names = TRUE)
head(mvc_nh)

mvc_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Age/H.xlsx", col_names = TRUE)
head(mvc_h) 

mvc_nh <- mvc_nh %>% 
  filter(Sex == "Male") %>% 
  ucd_age_clean_nh( "mvc")

mvc_h <- mvc_h %>% 
  filter(Sex == "Male") %>%
  ucd_age_clean_h( "mvc") 

age_mvc <-  bind_rows(
  mvc_nh,
  mvc_h
) 
head(age_mvc)

age_mvc <- race_age_format(age_mvc)
age_mvc %>% 
  group_by(race_eth) %>% 
  slice(1:3)

#adding population data
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarize(population = sum(population))
head(pop) 

age_mvc <- age_mvc %>%
  left_join(pop, by = c("race_eth", "age_group"))
head(age_mvc)
write.csv(age_mvc, "../CleanData/mvc_UCD_age.csv")
``` 

## By Year 

```{r}
#NH
mvc_y_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Year/NH.xlsx", col_names = TRUE)
head(mvc_y_nh)
mvc_y_nh <- ucd_year_clean_nh(mvc_y_nh, "mvc") 

#H
mvc_y_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Year/H.xlsx", col_names = TRUE)
head(mvc_y_h)
mvc_y_h <- ucd_year_clean_h(mvc_y_h, "mvc") 

#Binding
year_mvc <-  bind_rows(
  mvc_y_nh,
  mvc_y_h
) 


#Cleaning final merge

year_mvc <- year_mvc %>% 
  mutate(
    race_eth = case_when(
      race == "Black or African American" ~ "black",
      race == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    )
  )

glimpse(year_mvc) 
unique(year_mvc$race_eth) 

#adding population data 
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    age_group %in% c("15-19","20-24"),
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, year) %>% 
  summarize(population = sum(population))
head(pop) 

year_mvc <- year_mvc %>%
  left_join(pop, by = c("race_eth", "year"))
head(year_mvc)

write.csv(year_mvc, "../CleanData/mvc_UCD_yearly.csv")


```

# COD Males 15-24 2007-2020 

For our table 1 data, we need to make a larger dataframe that combines all the causes of death by race/ethnicity among just males aged 15-24 over the study period.  

We will start by pulling up the dataframes that already meet these conventions. We are also going to just retain cod, race_eth, and deaths.

```{r}
#mvc
age_mvc <- age_mvc %>% 
  filter(age_group %in% c("15-19", "20-24")) %>% 
  select(cod, race_eth, age_group, deaths) 
head(age_mvc)
#homicide
age_hom <- age_hom2 %>% 
  filter(age_group %in% c("15-19", "20-24")) %>% 
  select(cod, race_eth, age_group, deaths)
head(age_hom)
#all cause
head(all_cleaned) 
all_cause_table1 <- all_cleaned %>% 
  filter(year %in% c(2007:2020),
         sex == "male", 
         age_group %in% c("15-19", "20-24")) %>% 
  group_by(age_group, race_eth, cod) %>% 
  summarize(deaths = sum(deaths))
head(all_cause_table1)
``` 

Next we have to clean disease and injury accordingly: 
```{r} 

#function for later (import labeled the years strangely): 

#Disease 

#NH
dis_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Disease/Age/NH.xlsx", col_names = TRUE)
head(dis_nh)
dis_nh <- ucd_age_clean_nh(dis_nh, "disease")  


#H
dis_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Disease/Age/H.xlsx", col_names = TRUE)
head(dis_h)
dis_h <- ucd_age_clean_h(dis_h, "disease") 
 sum(is.na(dis_nh$Deaths) | dis_nh$Deaths == 0, na.rm = FALSE)
  

#Binding
age_dis <-  bind_rows(
 dis_h,
 dis_nh
) 

#Cleaning final merge
age_dis <- age_dis %>% 
  mutate(
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ),
    age_group = gsub(" years", "", age_group),
      age_group = trimws(age_group)
  )

head(age_dis)

# --------------------- 
#Injury  
#NH
inj_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Injuries/Age/NH.xlsx", col_names = TRUE)
head(inj_nh)
inj_nh <- ucd_age_clean_nh(inj_nh, "injury") 

#H
inj_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Injuries/Age/H.xlsx", col_names = TRUE)
head(inj_h)
inj_h <- ucd_age_clean_h(inj_h, "injury") 
 sum(is.na(inj_nh$Deaths) | inj_nh$Deaths == 0, na.rm = FALSE)

#Binding
age_inj <-  bind_rows(
 inj_h,
 inj_nh
) 

#Cleaning final merge
age_inj <- age_inj %>% 
  mutate(
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ),
    age_group = gsub(" years", "", age_group),
      age_group = trimws(age_group)
  )
  

head(age_inj) 

# --------------------- 
#Suicide  
#NH
sui_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Suicide/Age/NH.xlsx", col_names = TRUE)
head(sui_nh)
sui_nh  <- ucd_age_clean_nh(sui_nh , "suicide") 

#H
sui_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Suicide/Age/H.xlsx", col_names = TRUE)
head(sui_h)
sui_h <- ucd_age_clean_h(sui_h, "suicide") 
 sum(is.na(sui_h$Deaths) | sui_h$Deaths == 0, na.rm = FALSE)

#Binding
age_sui <-  bind_rows(
 sui_h,
 sui_nh
) 

#Cleaning final merge
age_sui <- age_sui %>% 
 mutate(
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ),
    age_group = gsub(" years", "", age_group),
      age_group = trimws(age_group)
  )
  

head(age_sui) 

``` 

Now that we have our datasets in the right configuration, we can bind all of them to make one large dataset with all causes of death among the population (15-24 year old males).

```{r}
table1 <- bind_rows(
  all_cause_table1,
  age_hom,
  age_mvc,
  age_inj,
  age_sui,
  age_dis
) 

head(table1) 
glimpse(table1) 

write_csv(table1, "../CleanData/table1.csv")
``` 


# Population Data 

Next, we are going to import the Population Estimates from 2001-2025 from the DOF. There are 4 separate datasets we will need to clean and merge (1990-1999,2001-2010, 2011-2020, 2021-2025). 

### 1990-1999 
For this dataset, we only need the year 1999 to supply population data for our entire study period. The data is not tidy, so we will need to write specific race_eth designations. 

```{r dof_99}
raw<- read_excel("../RawData/dof/Intercensal_1990_1999/California.xlsx", col_names = FALSE)
# Extract header rows
header1 <- raw[1, ]
header2 <- raw[2, ]

# Combine headers
colnames(raw) <- c(
  "area", "year", "sex", "age",
  paste(header1[5:ncol(raw)], header2[5:ncol(raw)], sep = "_")
)

# Drop header rows
dat <- raw %>% slice(-(1:2)) %>% clean_names()

head(dat)

#keeping Hispanic total
dat <- dat %>%
  mutate(
    across(
      starts_with(c("non_hispanic", "hispanic")),
      ~ as.numeric(.)
    )
  )

dof_99_tidy <- dat %>%
  filter(year == "1999") %>% 
  mutate(
    hispanic_total =
      hispanic_white +
      hispanic_black +
      hispanic_asian_pacific +
      hispanic_indian
  ) %>%
  select(
    -area, year, sex, age,
    non_hispanic_white,
    non_hispanic_black,
    hispanic_total
  ) %>%
  pivot_longer(
    cols = c(
      non_hispanic_white,
      non_hispanic_black,
      hispanic_total
    ),
    names_to = "race_eth",
    values_to = "population"
  ) %>%
  mutate(
    race_eth = case_when(
      race_eth == "non_hispanic_white" ~ "white",
      race_eth == "non_hispanic_black" ~ "black",
      race_eth == "hispanic_total"     ~ "hispanic"
    ),
    year = as.numeric(year),
    age = as.numeric(age), 
    sex = str_to_lower(sex)
  )

head(dof_99_tidy)


```


### 2000-2009  

This file contains both county-level and state-level population estimates. We will filter to only include the population estimates when county = 59, which corresponds with state-level estimates. We also will drop the estimate for the year 2010. After cleaning and cross-examining with the following dataframe, I found that the estimates were identical (Census-based) from this dataset and the 2010-2020 dataset. To ensure we don't inflate the estimate for 2010 by summing both datsets, we're just going to drop it from this one. 

We need to do some cleaning to make the variables consistent with other datasets. We will convert the Year to a proper "date" variable and select only the races/ethnic groups to be used in the study.

```{r dof1}
dof1 <- read.csv("../RawData/dof/Intercensal_2000-2010_DBInput_csv/Intercensal_2000-2010_DBInput_csv.txt") 
head(dof1) 

dof1_tidy <- dof1 %>%
  mutate(
    year = year(mdy_hms(Year)),
    age = as.numeric(Age),
    population = as.numeric(Population),
    race_eth = str_to_lower(RaceName),
    sex = str_to_lower(Gender)
  ) %>%
  filter(
    CountyCode == 59,
    race_eth %in% c("white", "black", "hispanic"),
    !Year %in% c("7/1/2000 0:00:00", "4/1/2010 0:00:00") #getting rid of duplicate 2000 and 2010 census
  ) %>%
  group_by(year, race_eth, sex, age) %>%
  summarise(population = sum(population), .groups = "drop") %>%
  arrange(year, race_eth, sex, age)

nrow(dof1_tidy) 
head(dof1_tidy)

``` 


### 2010-2020 

```{r dof2}
dof2 <- read_excel("../RawData/dof/Intercensal_Excel_California.xlsx",
                   sheet = "California-State",
                   col_names = TRUE) 
dof2 <- dof2 %>% 
  select(Sex:"April 1, 2020") %>% 
  rename(age = "Age (0-100+)",
         race_eth = "Race/ethnicity recode") %>% 
  rename_with(tolower)
 
```
Each row contains the population count for a given age, race, and sex. We are going to tidy this data by making a new "Year" variable and "Population" variable to contain the values that are in wide format right now. We also have two population values for 2010 (April, when the census took effect, then July). We will retain just the April counts, as they require no estimation and are the most accurate approximation of the population that year. 

```{r tidy_dof2}
dof2_tidy <- pivot_longer(dof2,
                     cols = 'april 1, 2010':'april 1, 2020',
                     names_to = "Year",
                     values_to = "Population")
dof2_tidy <- dof2_tidy %>% 
  filter(Year != "july 1, 2010") %>% 
  mutate(
    Year = year(mdy(Year)),
    sex = str_to_lower(sex)
  ) %>% 
   rename_with(tolower) %>% 
  filter(race_eth %in% c("White NH", "Black NH", "Hispanic")) %>% 
  mutate(race_eth = recode(
    race_eth, 
    "White NH" = "White", 
    "Black NH" = "Black"),
    race_eth = str_to_lower(race_eth),
    population = as.numeric(population)
  )
  
```
Now that our datasets have consistent tidy structure and naming conventions, we can stack them. 

```{r merge}
#Checking counts prior to merge
nrow(dof1_tidy) #6666
nrow(dof2_tidy) #6666
nrow(dof_99_tidy) #606

dof_data <- bind_rows(dof_99_tidy, dof1_tidy, dof2_tidy)
head(dof_data) 
nrow(dof_data) #13938, correct number of observations
```

Checking how the data may/may not change by different years.. 
```{r comparison_graphs}
dof1 <- dof1_tidy %>% mutate(source = "dof1")
dof2 <- dof2_tidy %>% mutate(source = "dof2")
dof99 <- dof_99_tidy %>% mutate(source = "dof99")

dof_all <- bind_rows(dof1, dof2, dof99)

pop_totals <- dof_all%>%
  group_by(source, year) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop")

pop_totals

ggplot(pop_totals, aes(x = year, y = population, color = source)) +
  geom_line(linewidth = 1) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Total population by year and data source",
    y = "Total population"
  )

#for each race/eth 
pop_race <- dof_all %>% 
  group_by(year, race_eth) %>% 
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop")


ggplot(pop_race, aes(year, population, color = race_eth))+
  geom_line(linewidth = 1) +
  geom_point() +
  theme_minimal() 
```


We will have two dataframes, one with the continuous ages so we may replicate one of the figures that has trends by each individual age. We will create a different dataframe that have age groups separately.   

```{r age_groups}
age_breaks <- c(seq(0, 90, by = 5), Inf)  # 0,5,10,...,90, Inf
age_labels <- c(paste(seq(0, 85, by = 5), seq(4, 89, by = 5), sep = "-"), "90+")

dof_data_grouped <- dof_data %>%
  mutate(age_group = cut(age, breaks = age_breaks, labels = age_labels, right = TRUE, include.lowest = TRUE)) %>%
  group_by(sex, race_eth, age_group, year) %>%
  summarise(population = sum(population, na.rm = TRUE), .groups = "drop") 
head(dof_data_grouped)  

dof_data_grouped %>% 
  filter(race_eth == "black") %>% 
  group_by(year, sex, age_group) %>% 
  summarize(n = n(), .groups = "drop") %>% 
  filter(n > 1) 

#how black population changes by age
dof_data_grouped %>% 
  filter(race_eth == "black") %>% 
  group_by(year, age_group) %>% 
  summarize(pop = sum(population, na.rm = TRUE)) %>% 
  ggplot(aes(year, pop, color = age_group)) +
  geom_line()

dof_data_grouped %>% 
  filter(race_eth == "black") %>% 
  group_by(year) %>% 
  summarize(pop = sum(population))
```
## 2021-2025

For 2021-2025, the only option (that I can ascertain) is to use the population projections for the years 2021-2025. We will be using the "Complete P-3 Race/Ethnicity and Sex by Age for California and Counties, Excel Compatible" California-State section from the [Projection webpage of the dof](https://dof.ca.gov/forecasting/demographics/projections/). 

```{r dof3}
dof3 <- read_excel("../RawData/dof/Projections/P3_California-and-Counties.xlsx",
                   sheet = "California-State",
                   col_names = TRUE) 
dof3 <- dof3 %>% 
  select(Sex:"2025", -"2020") %>% 
  rename(age = "Age (0-110+)",
         race_eth = "Race/ethnicity recode") %>% 
  rename_with(tolower) %>% 
  filter(race_eth %in% c(1,2,7)) %>% 
  mutate(sex = str_to_lower(sex),
         race_eth = recode(
    race_eth, 
    "1" = "White", 
    "2" = "Black",
    "7" = "Hispanic")) %>% 
  pivot_longer(
                     cols = '2021':'2025',
                     names_to = "year",
                     values_to = "population"
                     )
```

Checking that continuity is good. 

```{r} 

all_cleaned %>%
  filter(year %in% c(2016, 2017, 2018)) %>%
  group_by(year, race_eth) %>%
  summarise(deaths = sum(deaths)) 

  
```
#Saving population datasets 

```{r save}
write_csv(dof_data_grouped, "../CleanData/dof_data_grouped.csv") 
write_csv(dof_data, "../CleanData/dof_data.csv")
``` 
