---
title: "Analysis"
output_dir: ../Output 
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)
library(tidyr)

# Global options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

``` 

This document contains the analyses used to replicate Vaca et al. (2011). Datasets are pulled from the `/Cleandata` directory, which were produced in the `/scripts/Cleaning.Rmd` document.  

# Descriptive Analysis 

We will begin by exploring general descriptive statistics from our mortality data (2007-2020). 


```{r explore_population, include = FALSE}
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data,
                                by = c("year", "sex", "race_eth", "age_group"))
head(cdc_with_pop)
```
## Total Death Count 

How many deaths occurred over the study period?
```{r} 
mortality_07_20 <- cdc_with_pop %>% 
  filter(year %in% 2007:2020)

mortality_07_20 %>% 
  filter(cod == "all_cause") %>% 
  summarize(total_deaths = sum(deaths))

``` 

## Age-adjusted mortality rates 

What are the age-adjusted mortality rates for each race/ethnicity group? 

```{r} 

#Step 1: keep deaths and population by age group 
data_age <- cdc_with_pop %>%
  filter(
    year %in% 2007:2020, 
    cod == "all_cause"
  ) %>%
  group_by(race_eth, age_group) %>%
  summarise(deaths = sum(deaths),
            pop = sum(population), 
            .groups="drop")  
head(data_age)

# Step 2: Obtain standard population weight
std_pop <- dof_data %>%
  filter(
    year == 2000,
    sex == "male"
  ) %>%
  group_by(age_group) %>%
  summarise(std_pop = sum(population), .groups="drop") %>%
  mutate(weight = std_pop / sum(std_pop))

```


#Mortality rates per 100,000 

"Mortality rates per 100 000 population were computed for
Latinos, non-Latino whites, and non-Latino blacks by sex and by
5 year age groups (with ages 90 and older forming one age
group)." 

We will first calculate the all-cause mortality rates stratified by sex and age. 




#Overall comparisons 
"For the overall comparisons, mortality rates were
adjusted to the age distribution of the 2000 California Department of Finance population estimates." 

#Racial/ethnic mortality rate ratios 
"To compare Latino and
non-Latino black mortality to non-Latino white mortality, we
calculated mortality rate ratios. For both mortality rates and rate
ratios, confidence intervals (95% CI) were calculated using the
log normal approximation to the Poisson distribution. Trends
were assessed using variance weighted least squares regression"