---
title: "Analysis"
output_dir: ../Output 
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)
library(tidyr)

# Global options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=7)
source("../Scripts/functions.R")
``` 

This document contains the analyses used to replicate Vaca et al. (2011). Datasets are pulled from the `/Cleandata` directory, which were produced in the `/scripts/Cleaning.Rmd` document.  

# Descriptive Analysis 

We will begin by exploring general descriptive statistics from our mortality data (2007-2020). 


```{r explore_population, include = FALSE}
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data,
                                by = c("year", "sex", "race_eth", "age_group"))
head(cdc_with_pop) 

#-----------------------------------
# MERGE ONCE
#-----------------------------------

cdc <- cdc_mort %>%
  left_join(dof_data,
            by=c("year","sex","race_eth","age_group"))


yrs_0720 <- 2007:2020


```
## Total Death Count 

How many deaths occurred over the study period?
```{r} 
mortality_07_20 <- cdc_with_pop %>% 
  filter(year %in% 2007:2020)

mortality_07_20 %>% 
  filter(cod == "all_cause") %>% 
  summarize(total_deaths = sum(deaths))

``` 

## Age-adjusted mortality rates 

What are the age-adjusted mortality rates for each race/ethnicity group? 

```{r} 

#Step 1: keep deaths and population by age group 
data_age <- cdc_with_pop %>%
  filter(
    year %in% 2007:2020, 
    cod == "all_cause",
    !is.na(age_group)
  ) %>%
  group_by(race_eth, age_group) %>%
  summarise(deaths = sum(deaths),
            pop = sum(population), 
            .groups="drop")  
head(data_age)

# Step 2: Obtain standard population weight
std_pop <- dof_data %>%
  filter(
    year == 2000
  ) %>%
  group_by(age_group) %>%
  summarise(std_pop = sum(population), .groups="drop") %>%
  mutate(weight = std_pop / sum(std_pop))

# Step 3: compute age-specific rates
data_age <- data_age %>%
  mutate(rate = deaths / pop)
 
# Step 4: join weights + calculate weighted rates 
adj_rates <- data_age %>%
  left_join(std_pop, by="age_group") %>%
  mutate(weighted_rate = rate * weight)

#sum weighted rates --> age-adjusted rates
table1_adj <- adj_rates %>%
  group_by(race_eth) %>%
  summarise(
    age_adj_rate = sum(weighted_rate) * 100000,
    .groups="drop"
  )
table1_adj
```

# All cause mortality rates

## Fig 1- Male 
```{r}
fig1 <- cdc %>%
  calc_rates(yrs_0720,"male","all_cause") %>%
  calc_rr()

#plotting
plot_rr(fig1,NULL)

#range for 60+
fig1 %>% 
  filter(age_group %in% 
           c("60-64", "65-69", "70-74", "75-89", "90+")) %>% 
  group_by(race_eth, age_group) %>% 
  select(rr, rr_lcl, rr_ucl) 

#hispanic males 0-59 
fig1 %>% 
  filter(race_eth == "hispanic", 
         age_group %in% 
           c("0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59")) %>% 
  group_by(age_group) %>% 
  select(rr, rr_lcl, rr_ucl) %>% 
  arrange(rr)
```
Looking through the numbers: 

```{r}
tables_by_race <- fig1 %>%
  mutate(rr_ci = sprintf("%.2f (%.2f, %.2f)", rr, rr_lcl, rr_ucl)) %>%
  select(race_eth, age_group, rr_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(tables_by_race) <- unique(fig1$race_eth)

tables_by_race

```


## Fig 1- Female 

```{r}
fig2 <- cdc %>%
  calc_rates(yrs_0720,"female","all_cause") %>%
  calc_rr()

plot_rr(fig2, NULL)

```

Looking through the numbers: 

```{r}
tables_by_race <- fig2 %>%
  mutate(rr_ci = sprintf("%.2f (%.2f, %.2f)", rr, rr_lcl, rr_ucl)) %>%
  select(race_eth, age_group, rr_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(tables_by_race) <- unique(fig1$race_eth)

tables_by_race

``` 

# Homicide mortality rates

## Fig 3- Male 
```{r}
fig3 <- cdc %>%
  calc_rates(yrs_0720,"male","homicide") %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_rate(fig3,NULL)

``` 

By number: 
```{r}
tables_by_race <- fig3 %>%
  mutate(r_ci = sprintf("%.2f (%.2f, %.2f)", rate, lcl, ucl)) %>%
  select(race_eth, age_group, r_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(tables_by_race) <- unique(fig3$race_eth)

tables_by_race 
```

# All cause minus homicide 
## Figure 4: Male mortality rate ratios 

```{r}
fig4 <- cdc %>%
  filter(year %in% yrs_0720, sex=="male") %>%
  group_by(race_eth,age_group,cod) %>%
summarise(deaths=sum(deaths), .groups="drop") %>%
left_join(
  cdc %>%
    filter(year %in% yrs_0720, sex=="male") %>%
    group_by(race_eth,age_group) %>%
    summarise(pop=sum(population), .groups="drop"),
  by=c("race_eth","age_group")
) %>%
  pivot_wider(names_from=cod, values_from=deaths, values_fill=0) %>%
  mutate(deaths=all_cause-homicide,
         rate=deaths/pop*100000) %>%
  set_age_order() %>%
  calc_rr()

plot_rr(fig4,NULL)
``` 

Numerically: 
```{r}
table4_by_race <- fig4 %>%
  mutate(rr_ci = sprintf("%.2f (%.2f, %.2f)", rr, rr_lcl, rr_ucl)) %>%
  select(race_eth, age_group, rr_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(table4_by_race) <- unique(fig4$race_eth)

table4_by_race
``` 

# 
