---
title: "Analysis"
output_dir: ../Output 
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)
library(tidyr)

# Global options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

``` 

This document contains the analyses used to replicate Vaca et al. (2011). Datasets are pulled from the `/Cleandata` directory, which were produced in the `/scripts/Cleaning.Rmd` document.  

# Descriptive Analysis 

We will begin by exploring general descriptive statistics from our mortality data (2007-2020). 


```{r explore_population, include = FALSE}
dof_data_grouped <- read.csv("../CleanData/dof_data_grouped.csv")
dof_data <- read.csv("../CleanData/dof_data.csv")
```


#Mortality rates per 100,000 

"Mortality rates per 100 000 population were computed for
Latinos, non-Latino whites, and non-Latino blacks by sex and by
5 year age groups (with ages 90 and older forming one age
group)." 

We will first calculate the all-cause mortality rates stratified by sex and age. 




#Overall comparisons 
"For the overall comparisons, mortality rates were
adjusted to the age distribution of the 2000 California Department of Finance population estimates." 

#Racial/ethnic mortality rate ratios 
"To compare Latino and
non-Latino black mortality to non-Latino white mortality, we
calculated mortality rate ratios. For both mortality rates and rate
ratios, confidence intervals (95% CI) were calculated using the
log normal approximation to the Poisson distribution. Trends
were assessed using variance weighted least squares regression"