---
title: "Exploration"
output:
  html_document:
output_dir: ../Output
---
```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

```
# Exploring population data 

We are going to explore the distribution of our population data by individual ages and age categories, respectively. 

```{r explore_population, include = FALSE}
dof_data_grouped <- read.csv("../CleanData/dof_data_grouped.csv")
dof_data <- read.csv("../CleanData/dof_data.csv")
```

Let's examine the trends for the whole study period (data for 1999-2020):

```{r explore_age_groups, echo = FALSE} 
age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84", "85-89", "90+"
)
agg_data <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  group_by(race_eth, age_group) %>%
  summarise(total_population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  
  # convert to per 100,000
  mutate(pop_per_100k = total_population / 100000)

ggplot(agg_data, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()


``` 

Now let's look through the decades: 

```{r, echo = FALSE}
agg_period <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  mutate(period = case_when(
    year >= 2000 & year <= 2009 ~ "2000-2009",
    year >= 2010 & year <= 2020 ~ "2010-2020",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(race_eth, age_group, period) %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  ungroup() %>% 
   mutate(pop_per_100k = total_population / 100000)

ggplot(agg_period, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  facet_wrap(~period, ncol = 1) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()

```
# Sex differences 

```{r, include = FALSE}
population_table_wide <- dof_data_grouped %>%
  tidyr::pivot_wider(
    names_from = sex,
    values_from = population
  ) %>%
  arrange(age_group, year)

population_table_wide %>%
  kable(
    caption = "Population by Sex, Age Group, and Year"
  )

```


#All-cause mortality

To start, let's just look at all-cause mortality for males by racial/ethnic group. We are going to create a new dataframe for just this information: 
```{r all_cause_mort, echo = FALSE}
all_cause_mort <- read.csv("../CleanData/all_cause_mort.csv")

race_eth_all_cause <- all_cause_mort %>% 
  filter(Strata_Name %in% c("Female * Black", 
                            "Female * Hispanic", "Female * White",
                            "Male * Black", "Male * Hispanic",
                            "Male * White")) #excluding deaths for Multi-Race and Other/Unknown 

race_eth_all_cause <- race_eth_all_cause %>% #tidying
  separate(
    Strata_Name,
    into = c("sex", "race_eth"),
    sep = " \\* "
  ) %>% 
  select(-Strata) %>% 
  rename_with(tolower) %>% 
  mutate(sex = str_to_lower(sex),
         race_eth = str_to_lower(race_eth))
head(race_eth_all_cause)

```
Now that all of the naming conventions are the same as the population data, we can bring in the pop dataset and merge it with our race_eth_all_cause dataframe and calculate the death rate. We also need to apply the log-normal (Posson) CI for the rates.

```{r all_cause_with_pop, echo = FALSE}
dof_no_age <- read_csv("../CleanData/dof_no_age.csv")

all_cause_with_pop <- left_join(race_eth_all_cause,dof_no_age,
                                by = c("year", "sex", "race_eth"))

all_cause_with_pop <- all_cause_with_pop %>% 
  mutate( #mortality rate
          rate = count/population,
         rate_per_100k = rate*100000,
         
         # log-normal Poisson CI
    rate_lcl = rate * exp(-1.96 / sqrt(count)),
    rate_ucl = rate * exp( 1.96 / sqrt(count)),

    rate_lcl_100k = rate_lcl * 100000,
    rate_ucl_100k = rate_ucl * 100000)
head(all_cause_with_pop)
``` 


Let's summarize these trends: 

```{r all_c_no_age_trends, include = FALSE}
all_cause_table <- all_cause_with_pop %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_per_100k,
      rate_lcl_100k,
      rate_ucl_100k
    )
  )

#females
female_table <- all_cause_table %>%
  filter(sex == "female") %>%
  select(year, race_eth, rate_ci)

female_table %>%
  kable(
    caption = "ALl-Case Female Death Rate"
  )

#males 
male_table <- all_cause_table %>%
  filter(sex == "male") %>%
  select(year, race_eth, rate_ci)

male_table%>%
  kable(
    caption = "ALl-Case Male Death Rate"
  )
``` 

```{r echo = FALSE}
#sex stratified table
sex_stratified_table <- all_cause_table %>%
  select(year, sex, race_eth, rate_ci) %>%
  pivot_wider(
    names_from = sex,
    values_from = rate_ci
  )

sex_stratified_table%>%
  kable(
    caption = "ALl-Case Death Rate Stratified by Sex"
  )
```

*Mortality rates are expressed per 100,000 persons. 95% confidence intervals were calculated using the log-normal approximation to the Poisson distribution.*

Now, we are going to plot. We are going to use White as the referent category. 

```{r plot_no_age_rr, echo = FALSE}
rr_data <- all_cause_with_pop %>%
  group_by(year, sex) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # drop reference from RR plot
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1/count + 1/ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1/count + 1/ref_count))
  )
rr_data
#females
ggplot(
  rr_data %>% filter(sex == "female"),
  aes(x = year, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Females",
    subtitle = "Reference group: White",
    x = "Year",
    y = "Rate Ratio"
  ) +
  theme_minimal()

#males 
ggplot(
  rr_data %>% filter(sex == "male"),
  aes(x = year, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Males",
    subtitle = "Reference group: White",
    x = "Year",
    y = "Rate Ratio"
  ) +
  theme_minimal()

```

# CDC Mortality Data 

```{r cdc_merge, include = FALSE}
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
head(cdc_mort)

#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")
head(dof_data)

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data,
                                by = c("year", "sex", "race_eth", "age_group"))

head(cdc_with_pop)
```


# Replicating tables and figures 
## 2006-2023 data 


