---
title: "Exploration"
output:
  html_document:
output_dir: ../Output
---
```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)
library(tidyr)

# Global options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

```
# Exploring population data 

We are going to explore the distribution of our population data by individual ages and age categories, respectively. 

```{r explore_population, include = FALSE}
dof_data_grouped <- read.csv("../CleanData/dof_data_grouped.csv")
dof_data <- read.csv("../CleanData/dof_data.csv")
```

Let's examine the trends for the whole study period (data for 1999-2020):

```{r explore_age_groups, echo = FALSE} 
age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84", "85-89", "90+"
)
agg_data <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  group_by(race_eth, age_group) %>%
  summarise(total_population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  
  # convert to per 100,000
  mutate(pop_per_100k = total_population / 100000)

ggplot(agg_data, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()


``` 

Now let's look through the decades: 

```{r, echo = FALSE}
agg_period <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  mutate(period = case_when(
    year >= 2000 & year <= 2009 ~ "2000-2009",
    year >= 2010 & year <= 2020 ~ "2010-2020",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(race_eth, age_group, period) %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  ungroup() %>% 
   mutate(pop_per_100k = total_population / 100000)

ggplot(agg_period, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  facet_wrap(~period, ncol = 1) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()

```
# Sex differences 

```{r, include = FALSE}
population_table_wide <- dof_data_grouped %>%
  tidyr::pivot_wider(
    names_from = sex,
    values_from = population
  ) %>%
  arrange(age_group, year)

population_table_wide %>%
  kable(
    caption = "Population by Sex, Age Group, and Year"
  )

```


#All-cause mortality

To start, let's just look at all-cause mortality for males by racial/ethnic group. We are going to create a new dataframe for just this information: 
```{r all_cause_mort, echo = FALSE}
all_cause_mort <- read.csv("../CleanData/all_cause_mort.csv")

race_eth_all_cause <- all_cause_mort %>% 
  filter(Strata_Name %in% c("Female * Black", 
                            "Female * Hispanic", "Female * White",
                            "Male * Black", "Male * Hispanic",
                            "Male * White")) #excluding deaths for Multi-Race and Other/Unknown 

race_eth_all_cause <- race_eth_all_cause %>% #tidying
  separate(
    Strata_Name,
    into = c("sex", "race_eth"),
    sep = " \\* "
  ) %>% 
  select(-Strata) %>% 
  rename_with(tolower) %>% 
  mutate(sex = str_to_lower(sex),
         race_eth = str_to_lower(race_eth))
head(race_eth_all_cause)

```
Now that all of the naming conventions are the same as the population data, we can bring in the pop dataset and merge it with our race_eth_all_cause dataframe and calculate the death rate. We also need to apply the log-normal (Posson) CI for the rates.

```{r all_cause_with_pop, echo = FALSE}
dof_no_age <- read_csv("../CleanData/dof_no_age.csv")

all_cause_with_pop <- left_join(race_eth_all_cause,dof_no_age,
                                by = c("year", "sex", "race_eth"))

all_cause_with_pop <- all_cause_with_pop %>% 
  mutate( #mortality rate
          rate = count/population,
         rate_per_100k = rate*100000,
         
         # log-normal Poisson CI
    rate_lcl = rate * exp(-1.96 / sqrt(count)),
    rate_ucl = rate * exp( 1.96 / sqrt(count)),

    rate_lcl_100k = rate_lcl * 100000,
    rate_ucl_100k = rate_ucl * 100000)
head(all_cause_with_pop)
``` 


Let's summarize these trends: 

```{r all_c_no_age_trends, include = FALSE}
all_cause_table <- all_cause_with_pop %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_per_100k,
      rate_lcl_100k,
      rate_ucl_100k
    )
  )

#females
female_table <- all_cause_table %>%
  filter(sex == "female") %>%
  select(year, race_eth, rate_ci)

female_table %>%
  kable(
    caption = "ALl-Case Female Death Rate"
  )

#males 
male_table <- all_cause_table %>%
  filter(sex == "male") %>%
  select(year, race_eth, rate_ci)

male_table%>%
  kable(
    caption = "ALl-Case Male Death Rate"
  )
``` 

```{r echo = FALSE}
#sex stratified table
sex_stratified_table <- all_cause_table %>%
  select(year, sex, race_eth, rate_ci) %>%
  pivot_wider(
    names_from = sex,
    values_from = rate_ci
  )

sex_stratified_table%>%
  kable(
    caption = "ALl-Case Death Rate Stratified by Sex"
  )
```

*Mortality rates are expressed per 100,000 persons. 95% confidence intervals were calculated using the log-normal approximation to the Poisson distribution.*

Now, we are going to plot. We are going to use White as the referent category. 

```{r plot_no_age_rr, echo = FALSE}
rr_data <- all_cause_with_pop %>%
  group_by(year, sex) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # drop reference from RR plot
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1/count + 1/ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1/count + 1/ref_count))
  )
rr_data
#females
ggplot(
  rr_data %>% filter(sex == "female"),
  aes(x = year, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Females",
    subtitle = "Reference group: White",
    x = "Year",
    y = "Rate Ratio"
  ) +
  theme_minimal()

#males 
ggplot(
  rr_data %>% filter(sex == "male"),
  aes(x = year, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Males",
    subtitle = "Reference group: White",
    x = "Year",
    y = "Rate Ratio"
  ) +
  theme_minimal()

```

# CDC Mortality Data 

```{r cdc_merge, include = FALSE}
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
head(cdc_mort)

#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")
head(dof_data)

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data,
                                by = c("year", "sex", "race_eth", "age_group"))

head(cdc_with_pop)
```


# Replicating tables and figures 
## 2006-2023 data 

To start, I am going to exclude the data from 2023, as we will need to figure out whether we are going to make two separate tables (standardizing to 2010 and 2020 population, with understanding that years after 2020 come from population *projections*, not estimates) or just include 2006-2020 data.

## Table 1

To replicate the table, we need to:  

1. subset deaths to males 15-24, years 2006-2020

2. create a population weighting variable for 2010

3. compute age-specific weights 

4. Join weights to calculate age-adjusted rates 

5. Calculate 95% CIs using log normal approximation to the Poisson distribution 


```{r}

#Step 1: keep deaths and population by age group 
data_age <- cdc_mort %>%
  filter(
    year %in% 2006:2020,
    age_group %in% c("15-19","20-24"),
    sex=="male"
  ) %>%
  group_by(race_eth, cod, age_group) %>%
  summarise(deaths = sum(deaths), .groups="drop") %>%
  left_join(
    dof_data_grouped %>%
      filter(
        year %in% 2006:2020,
        age_group %in% c("15-19","20-24"),
        sex=="male"
      ) %>%
      group_by(race_eth, age_group) %>%
      summarise(pop = sum(population), .groups="drop"),
    by=c("race_eth","age_group")
  )

#Step 2: get standard population weights 
std_pop <- dof_data_grouped %>%
  filter(
    year == 2010,
    age_group %in% c("15-19","20-24"),
    sex == "male"
  ) %>%
  group_by(age_group) %>%
  summarise(std_pop = sum(population), .groups="drop") %>%
  mutate(weight = std_pop / sum(std_pop))

# Step 3: compute age-specific rates
data_age <- data_age %>%
  mutate(rate = deaths / pop)
 
# Step 4: join weights + calculate weighted rates 
adj_rates <- data_age %>%
  left_join(std_pop, by="age_group") %>%
  mutate(weighted_rate = rate * weight)

#sum weighted rates --> age-adjusted rates
table1_adj <- adj_rates %>%
  group_by(race_eth, cod) %>%
  summarise(
    age_adj_rate = sum(weighted_rate) * 100000,
    .groups="drop"
  )

# Step 5: adding poisson CI 
table1_adj <- adj_rates %>%
  group_by(race_eth, cod) %>%
  summarise(
    adj_rate = sum(weighted_rate),
    deaths_total = sum(deaths),
    .groups="drop"
  ) %>%
  mutate(
    rate_100k = adj_rate * 100000,

    se_log = 1 / sqrt(deaths_total),

    lcl = rate_100k * exp(-1.96 * se_log),
    ucl = rate_100k * exp( 1.96 * se_log)
  )

#formatting
table1_adj <- table1_adj %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_100k, lcl, ucl
    )
  )
cod_levels <- c(
  "homicide",
  "mvc",
  "suicide",
  "other_injury",
  "other_disease",
  "all_cause"
)

cod_labels <- c(
  homicide = "Homicide",
  mvc = "Motor Vehicle Crashes",
  suicide = "Suicide",
  other_injury = "Other Injuries",
  other_disease = "Other Disease",
  all_cause = "Total (All Causes)"
)

table1_final <- table1_final %>%
  mutate(
    cod = factor(cod, levels = cod_levels, ordered = TRUE),
    cod = recode(cod, !!!cod_labels)
  ) %>%
  arrange(cod)

knitr::kable(
  table1_final,
  caption = "Table 1. Annual mortality rates (95% CI) per 100,000 by cause of death and race/ethnic group. California males age 15–24 years, 2006–2020."
)

``` 
We are going to use some of the functions defined in our functions.R script to conduct our exploration.

```{r figures}

source("../Scripts/functions.R")

#-----------------------------------
# MERGE ONCE
#-----------------------------------

cdc <- cdc_mort %>%
  left_join(dof_data_grouped,
            by=c("year","sex","race_eth","age_group"))


yrs_9906 <- 1999:2006
yrs_0620 <- 2006:2020



#================================================
# FIGURE 1 — Male all-cause RR
#================================================
fig1 <- cdc %>%
  calc_rates(yrs_0620,"male","all_cause") %>%
  calc_rr()

plot_rr(fig1,"Male All-Cause Mortality Rate Ratios")



#================================================
# FIGURE 2 — Female all-cause RR
#================================================
fig2 <- cdc %>%
  calc_rates(yrs_0620,"female","all_cause") %>%
  calc_rr()

plot_rr(fig2,"Female All-Cause Mortality Rate Ratios")



#================================================
# FIGURE 3 — Male homicide rates
#================================================
fig3 <- cdc %>%
  calc_rates(yrs_0620,"male","homicide") %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_rate(fig3,"Male Homicide Mortality Rates")



#================================================
# FIGURE 4 — Male all-cause minus homicide RR
#================================================
fig4 <- cdc %>%
  filter(year %in% yrs_9906, sex=="male") %>%
  group_by(race_eth,age_group,cod) %>%
summarise(deaths=sum(deaths), .groups="drop") %>%
left_join(
  cdc %>%
    filter(year %in% yrs_9906, sex=="male") %>%
    group_by(race_eth,age_group) %>%
    summarise(pop=sum(population), .groups="drop"),
  by=c("race_eth","age_group")
) %>%
  pivot_wider(names_from=cod, values_from=deaths, values_fill=0) %>%
  mutate(deaths=all_cause-homicide,
         rate=deaths/pop*100000) %>%
  set_age_order() %>%
  calc_rr()

plot_rr(fig4,"Male Non-Homicide Mortality RR")



#================================================
# FIGURE 5 — Homicide yearly (15–24 males)
#================================================
fig5 <- cdc %>%
  filter(sex=="male",
         age_group %in% c("15-19","20-24"),
         cod=="homicide",
         year<=2006) %>%
  group_by(year,race_eth) %>%
  summarise(deaths=sum(deaths),pop=sum(population),.groups="drop") %>%
  mutate(rate=deaths/pop*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_yearly(fig5,"Homicide Mortality Males 15–24")



#================================================
# FIGURE 6 — MVC rates by age
#================================================
fig6 <- cdc %>%
  calc_rates(yrs_9906,"male","mvc") %>%
  rowwise() %>%
  mutate(ci_log(rate,deaths)) %>%
  ungroup()

plot_rate(fig6,"Motor Vehicle Crash Mortality Rates")



#================================================
# FIGURE 7 — MVC yearly (15–24)
#================================================
fig7 <- cdc %>%
  filter(sex=="male",
         age_group %in% c("15-19","20-24"),
         cod=="mvc",
         year<=2006) %>%
  group_by(year,race_eth) %>%
  summarise(deaths=sum(deaths),pop=sum(population),.groups="drop") %>%
  mutate(rate=deaths/pop*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_yearly(fig7,"MVC Mortality Males 15–24")

```



## Figure 1 
 
We will need to 

**1. subset to males, all-cause only, retaining age groups, summing deaths for years 2006-2020**

**2. use population data from 2006-2020 to calculate overall rates by age groups**

**3. Calculate rate ratios with NH white as reference group ** 
For each 5-year age group, we first calculate all-cause mortality rates for each race/ethnicity as the total number of deaths from 1999–2006 divided by the corresponding summed population for the same period. Rates are expressed per 100,000 population.
Within each age group, we then calculate mortality rate ratios (RRs) by dividing the mortality rate for Latino males and non-Latino Black males by the mortality rate for non-Latino White males (reference group).
Ninety-five percent confidence intervals for the rate ratios are estimated using a Poisson approximation based on the number of deaths in the comparison and reference groups.

```{r figure_1}
#reordering ages
age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84",
  "85-89", "90+"
)

dof_data_grouped <- dof_data_grouped %>% 
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  ) 

#Step 1 
figure1_data <- cdc_mort %>% 
  filter(
    year %in% 2006:2020,
    sex == "male", 
    cod == "all_cause"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    death_count = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )


#Step 2
figure1_pop <- dof_data_grouped %>% 
  filter(
    year %in% 2006:2020,
    sex == "male"
  ) %>% 
  group_by(race_eth, age_group) %>%
  summarise(
    pop_2006_2020 = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) 


# Step 3:

figure1_rates <- figure1_data %>%
  left_join(
    figure1_pop,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    rate = (death_count / pop_2006_2020) * 100000
  )

# Calculate rate ratios using NH White as reference within age group
rr_data <- figure1_rates %>%
  group_by(age_group) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = death_count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # exclude reference group from plotting
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1 / death_count + 1 / ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1 / death_count + 1 / ref_count))
  )


rr_data <- rr_data %>%
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )

#plotting
ggplot(
  rr_data,
  aes(x = age_group, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Males 2006-2020",
    subtitle = "Reference group: White",
    x = "Age (years)",
    y = "Rate Ratio"
  ) +
  theme_minimal()

```


## Figure 2 

Next, we are going to replicate Figure 2, which is "Female all cause mortality rate ratios comparing Latino and non-Latino black mortality to non-Latino white mortality, by 5 year age groups with 95% CI, California, 2006-2020. 

We will need to replicate the previous steps, except subsetting to **females** instead of males. 

**1. Subset to females, all-cause only, retaining age groups, summing deaths for years 2006-2020**

**2. Use population data from 2006-2020 to calculate overall rates by age groups**

**3. Calculate rate ratios with NH white as reference group ** 
For each 5-year age group, we first calculate all-cause mortality rates for each race/ethnicity as the total number of deaths from 1999–2006 divided by the corresponding summed population for the same period. Rates are expressed per 100,000 population.
Within each age group, we then calculate mortality rate ratios (RRs) by dividing the mortality rate for Latino males and non-Latino Black males by the mortality rate for non-Latino White males (reference group).
Ninety-five percent confidence intervals for the rate ratios are estimated using a Poisson approximation based on the number of deaths in the comparison and reference groups.

```{r figure_2}
#reordering ages
age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84",
  "85-89", "90+"
)

dof_data_grouped <- dof_data_grouped %>% 
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  ) 

#Step 1 
figure2_data <- cdc_mort %>% 
  filter(
    year %in% 2006:2020,
    sex == "female", 
    cod == "all_cause"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    death_count = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )

#Step 2
figure2_pop <- dof_data_grouped %>% 
  filter(
    year %in% 2006:2020,
    sex == "female"
  ) %>% 
  group_by(race_eth, age_group) %>%
  summarise(
    pop_2006_2020 = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) 


# Step 3: merge deaths + population, calculate rates, then rate ratios

figure2_rates <- figure2_data %>%
  left_join(
    figure2_pop,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    rate = (death_count / pop_2006_2020) * 100000
  )

# Calculate rate ratios using NH White as reference within age group
rr_data_2 <- figure2_rates %>%
  group_by(age_group) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = death_count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # exclude reference group from plotting
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1 / death_count + 1 / ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1 / death_count + 1 / ref_count))
  )

rr_data_2 <- rr_data_2 %>%
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )

#plotting
ggplot(
  rr_data_2,
  aes(x = age_group, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Females 2006-2020",
    subtitle = "Reference group: White",
    x = "Age (years)",
    y = "Rate Ratio"
  ) +
  theme_minimal()

``` 

## Figure 3 

Next, we are going to replicate Figure 3, which is "Male homicide mortality rate ratios comparing Latino and non-Latino black mortality to non-Latino white mortality, by 5 year age groups with 95% CI, California, 2006-2020 (Overlapping 95% CI for age group 80 years and older not shown.)". 

We will start by selecting just homicide data, then replicate previous steps.

**1. Subset to females, all-cause only, retaining age groups, summing deaths for years 2006-2020**

**2. Use population data from 2006-2020 to calculate overall rates by age groups**

**3. Calculate death rates ** 

```{r figure_3}

#Step 1 
figure3_data <- cdc_mort %>% 
  filter(
    year %in% 2006:2020,
    sex == "male", 
    cod == "homicide"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    death_count = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  ) 


#Step 2
figure3_pop <- dof_data_grouped %>% 
  filter(
    year %in% 2006:2020,
    sex == "male"
  ) %>% 
  group_by(race_eth, age_group) %>%
  summarise(
    pop_2006_2020= sum(population, na.rm = TRUE),
    .groups = "drop"
  ) 


# Step 3: merge deaths + population, calculate rates, then 95% CI
figure3_rates <- figure3_data %>%
  left_join(
    figure3_pop,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    rate = (death_count / pop_2006_2020) * 100000,
    rate_lcl = rate * exp(-1.96 * sqrt(1 / death_count)),
    rate_ucl = rate * exp( 1.96 * sqrt(1 / death_count))
  ) %>%
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )


#plotting
ggplot(
  figure3_rates,
  aes(x = age_group, y = rate, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rate_lcl, ymax = rate_ucl),
    width = 0.15
  ) +
  labs(
    title = "Male homicide mortality rate ratios by Race/Ethnicity",
    subtitle = "1999 - 2006",
    x = "Age (years)",
    y = "Rate Ratio"
  ) +
  theme_minimal() 
```

## Figure 4

Next, we are going to replicate Figure 4, which is "Male mortality rate ratios for all causes except homicide, comparing Latino and non-Latino black mortality to non-Latino white mortality, by 5 year age groups with 95% CI, California, 1999-2006". 

We will start by calculating all cause - homicides, then replicate previous steps.

**1. Calculate all-case deaths - homicides**

**2. Subset to male, subtract homicides from all cause, retaining age groups, summing deaths for years 1999:2006**

**3. Use population data from 1999-2006 to calculate overall rates by age groups**

**4. Calculate rate ratios with NH white as reference group **  

```{r figure_4}
#Step 1  
#all-cause deaths
all_cause <- cdc_mort %>% 
  filter(
    year %in% 1999:2006,
    sex == "male",
    cod == "all_cause"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    all_deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )
#homicides 
homicide <- cdc_mort %>% 
  filter(
    year %in% 1999:2006,
    sex == "male",
    cod == "homicide"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    homicide_deaths = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )
#subtracting 
figure4_data <- all_cause %>%
  left_join(
    homicide,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    homicide_deaths = replace_na(homicide_deaths, 0),
    death_count = all_deaths - homicide_deaths
  ) %>%
  select(race_eth, age_group, death_count) %>% 
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )



#Step 2
figure4_pop <- dof_data_grouped %>% 
  filter(
    year %in% 1999:2006,
    sex == "male"
  ) %>% 
  group_by(race_eth, age_group) %>%
  summarise(
    pop_1999_2006 = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) 


# Step 3: merge deaths + population, calculate rates, then rate ratios

figure4_rates <- figure4_data %>%
  left_join(
    figure4_pop,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    rate = (death_count / pop_1999_2006) * 100000
  )

# Calculate rate ratios using NH White as reference within age group
rr4_data <- figure4_rates %>%
  group_by(age_group) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = death_count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # exclude reference group from plotting
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1 / death_count + 1 / ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1 / death_count + 1 / ref_count))
  )



rr4_data <- rr4_data %>%
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )

#plotting
ggplot(
  rr4_data,
  aes(x = age_group, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios all causes except homicide by Race/Ethnicity — Males",
    subtitle = "Reference group: White",
    x = "Age (years)",
    y = "Rate Ratio"
  ) +
  theme_minimal()


```


## Figure 5  

Next, we are going to replicate Figure 5, which is "Homicide mortality rates, among males 15-24 years of age, comparing Latino, non-Latino black, and non-Latino white, by year with 95% CI, California 1999-2006". 

We will start by selecting homicides among males 15-24, then replicate previous steps.

**1. Subset to males aged 15-24, homicides only, summing deaths for years 1999:2006**

**2. Use population data from 1999-2006 to calculate overall rates by age groups**

**3. Calculate death rates **


```{r figure_5, echo = FALSE, fig.width=9, fig.height= 4}
fig5_data <- cdc_with_pop %>%
  filter(
    sex == "male",
    age_group %in% c("15-19", "20-24"),
    cod == "homicide",
    year <= 2006
  )

fig5_yearly <- fig5_data %>%
  group_by(year, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )

fig5_yearly <- fig5_yearly %>%
  mutate(
    rate = deaths / population,
    rate_per_100k = rate * 100000
  )

fig5_yearly <- fig5_yearly %>%
  mutate(
    rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population * 100000,
    rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population * 100000
  )


ggplot(fig5_yearly,
       aes(x = year, y = rate_per_100k, color = race_eth)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rate_lcl, ymax = rate_ucl),
    width = 0.2
  ) +
  scale_x_continuous(breaks = 1999:2006) +
  scale_y_continuous(
    name = "Mortality rate per 100,000"
  ) +
  scale_color_manual(
    values = c(
      "hispanic" = "#E69F00",
      "black" = "#000000",
      "white" = "#56B4E9"
    ),
    labels = c(
      "hispanic" = "Latino",
      "black" = "Non-Latino Black",
      "white" = "Non-Latino White"
    )
  ) +
  labs(
    title= "Homicide Mortality Rate, Males 15-24 y/o, 2000-2006",
    x = "Year",
    color = "Race/Ethnicity"
  ) +
 theme_minimal()

```

## Figure 6 

This figure is comparing death rates for motor vehicle crashes by age groups ("Male motor vehicle crash mortality rates, comparing Latino, non-Latino black, and non-Latino white, 5 year age groups with 95% CI, California 1999-2006. (Points are offset to show CI; upper CI above 50/100,000 not shown).") 


```{r figure_6, echo = FALSE, fig.width=9, fig.height= 4}
#Step 1 
figure6_data <- cdc_mort %>% 
  filter(
    year %in% 1999:2006,
    sex == "male", 
    cod == "mvc"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    death_count = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )

#Step 2
figure6_pop <- dof_data_grouped %>% 
  filter(
    year %in% 1999:2006,
    sex == "male"
  ) %>% 
  group_by(race_eth, age_group) %>%
  summarise(
    pop_1999_2006 = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) 


# Step 3: merge deaths + population, calculate rates, then rate ratios

figure6_rates <- figure6_data %>%
  left_join(
    figure6_pop,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    rate = (death_count / pop_1999_2006) * 100000,
    rate_lcl = rate * exp(-1.96 * sqrt(1 / death_count)),
    rate_ucl = rate * exp( 1.96 * sqrt(1 / death_count))
  ) %>%
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )


#plotting
ggplot(
  figure6_rates,
  aes(x = age_group, y = rate, color = race_eth, group = race_eth)
) +
  geom_point(size = 2, position = position_dodge(width = 0.3)) +
  geom_line(position = position_dodge(width = 0.3)) +
  geom_errorbar(
    aes(ymin = rate_lcl, ymax = rate_ucl),
    width = 0.15,
    position = position_dodge(width = 0.3)
  ) +
  coord_cartesian(ylim = c(0, 50)) +
  labs(
    title = "Motor Vehicle Crash Mortality Rates by Race/Ethnicity — Males",
    subtitle = "California, 1999–2006",
    x = "Age group (years)",
    y = "Deaths per 100,000 population"
  ) +
  theme_minimal()
``` 

As you may notice, we are missing data for some age groups, more prominently surrounding NH Blacks that are less than 15 years old and over 60 years old. I ran the following check, but it appears that similar to the other plot, we just don't have mvc data for these age groups. We likely observe this because our cdc query just doesn't populate due to insufficient cell size. However, this is a bit surprising since MVC deaths appear pretty large among Non-Latino Blacks in Vaca et al (2011). One thing we might try is importing another dataset from cdc wonder, this time stratifying by 10 year age groups. For now, we will continue on... 
```{r checking_black_mvcs}
cdc_mort %>% 
  filter(cod == "mvc" 
         & sex == "male" 
         & age_group %in% c("60-64", "65-69")
         & race_eth == "black"
         & year %in% 1999:2006
           )
```

## Figure 7 

This figure also displays motor vehicle crash mortality rates, this time only including ages 15-24, comparing the rates for each race/ethnicity over each year in the study period. 

```{r figure_7, echo = FALSE}
fig7_data <- cdc_with_pop %>%
  filter(
    sex == "male",
    age_group %in% c("15-19", "20-24"),
    cod == "mvc",
    year <= 2006
  )

fig7_yearly <- fig7_data %>%
  group_by(year, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )

fig7_yearly <- fig7_yearly %>%
  mutate(
    rate = deaths / population,
    rate_per_100k = rate * 100000
  )

fig7_yearly <- fig7_yearly %>%
  mutate(
    rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population * 100000,
    rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population * 100000
  )

fig7_yearly %>% arrange(desc(rate_per_100k))


pd <- position_dodge(width = 0.3)

ggplot(fig7_yearly, aes(x = year, y = rate_per_100k, color = race_eth)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, position = pd) +
  geom_errorbar(
    aes(ymin = rate_lcl, ymax = rate_ucl),
    width = 0.2,
    position = pd
  ) +
  scale_x_continuous(breaks = 1999:2006) +
  scale_y_continuous(
    name = "Mortality rate per 100,000",
    limits = c(0, 50),
    breaks = seq(0, 50, by = 10)
  ) +
  labs(
    title = "Motor Vehicle Crash Mortality Rates, Males 15–24",
    x = "Year",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()


```



We are going to do the same thing, except this time we will extend the data to go through years 2006-2021:

```{r figure_7_extended, echo = FALSE, fig.width=11, fig.height= 6}
fig7x_data <- cdc_with_pop %>%
  filter(
    sex == "male",
    age_group %in% c("15-19", "20-24"),
    cod == "mvc",
    year >= 2006, 
    year < 2021
  )

fig7x_yearly <- fig7x_data %>%
  group_by(year, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )

fig7x_yearly <- fig7x_yearly %>%
  mutate(
    rate = deaths / population,
    rate_per_100k = rate * 100000
  )

fig7x_yearly <- fig7x_yearly %>%
  mutate(
    rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population * 100000,
    rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population * 100000
  )

fig7x_yearly %>% arrange(desc(rate_per_100k))


pd <- position_dodge(width = 0.3)

ggplot(fig7x_yearly, aes(x = year, y = rate_per_100k, color = race_eth)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, position = pd) +
  geom_errorbar(aes(ymin = rate_lcl, ymax = rate_ucl), width = 0.2, position = pd) +
  scale_x_continuous(breaks = 2006:2023) +
  scale_y_continuous(name = "Mortality rate per 100,000",
                     limits = c(0, 60) ) +
  
  labs(
    title = "Motor vehicle crash mortality rates males 15-24",
    x = "Year",
    color = "Race/Ethnicity",
    linetype = "Race/Ethnicity"
  ) +
  theme_minimal()
``` 

```{r fig7x_table, include = FALSE}
fig7x_table <- fig7x_yearly %>%
  select(year, race_eth, rate_per_100k) %>%
  pivot_wider(
    names_from = race_eth,       # new columns from race_eth
    values_from = rate_per_100k  # values to fill in
  )

fig7x_table
```
