---
title: "Exploration"
output:
  html_document:
output_dir: ../Output
---
```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)
library(tidyr)

# Global options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

```
# Exploring population data 

We are going to explore the distribution of our population data by individual ages and age categories, respectively. 

```{r explore_population, include = FALSE}
dof_data_grouped <- read.csv("../CleanData/dof_data_grouped.csv")
dof_data <- read.csv("../CleanData/dof_data.csv")
```


```{r explore_age_groups, include = FALSE} 
#Let's examine the trends for the whole study period (data for 1999-2020):
age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84", "85-89", "90+"
)
agg_data <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  group_by(race_eth, age_group) %>%
  summarise(total_population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  
  # convert to per 100,000
  mutate(pop_per_100k = total_population / 100000)

ggplot(agg_data, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()


``` 


```{r, include = FALSE}
#Now let's look through the decades: 
agg_period <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  mutate(period = case_when(
    year >= 2000 & year <= 2009 ~ "2000-2009",
    year >= 2010 & year <= 2020 ~ "2010-2020",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(race_eth, age_group, period) %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  ungroup() %>% 
   mutate(pop_per_100k = total_population / 100000)

ggplot(agg_period, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  facet_wrap(~period, ncol = 1) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()

```


```{r, include = FALSE}
# Sex differences 
population_table_wide <- dof_data_grouped %>%
  tidyr::pivot_wider(
    names_from = sex,
    values_from = population
  ) %>%
  arrange(age_group, year)

population_table_wide %>%
  kable(
    caption = "Population by Sex, Age Group, and Year"
  )

```



# CDC Mortality Data 

```{r cdc_merge, include = FALSE}
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
head(cdc_mort)

#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")
head(dof_data)

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data,
                                by = c("year", "sex", "race_eth", "age_group"))

head(cdc_with_pop)
```


# Replicating tables and figures 
## 2007-2023 data 

To start, I am going to exclude the data from 2023, as we will need to figure out whether we are going to make two separate tables (standardizing to 2010 and 2020 population, with understanding that years after 2020 come from population *projections*, not estimates) or just include 2006-2020 data.

## Table 1

To replicate the table, we need to:  

1. subset deaths to males 15-24, years 2007-2020

2. create a population weighting variable for 2000 to standardize age-adjustments

3. compute age-specific weights 

4. Join weights to calculate age-adjusted rates 

5. Calculate 95% CIs using log normal approximation to the Poisson distribution 


```{r}

#Step 1: keep deaths and population by age group 
data_age <- cdc_mort %>%
  filter(
    year %in% 2007:2020,
    age_group %in% c("15-19","20-24"),
    sex=="male"
  ) %>%
  group_by(race_eth, cod, age_group) %>%
  summarise(deaths = sum(deaths), .groups="drop") %>%
  left_join(
    dof_data_grouped %>%
      filter(
        year %in% 2007:2020,
        age_group %in% c("15-19","20-24"),
        sex=="male"
      ) %>%
      group_by(race_eth, age_group) %>%
      summarise(pop = sum(population), .groups="drop"),
    by=c("race_eth","age_group")
  )

#Step 2: get standard population weights 
std_pop <- dof_data_grouped %>%
  filter(
    year == 2000,
    age_group %in% c("15-19","20-24"),
    sex == "male"
  ) %>%
  group_by(age_group) %>%
  summarise(std_pop = sum(population), .groups="drop") %>%
  mutate(weight = std_pop / sum(std_pop))

# Step 3: compute age-specific rates
data_age <- data_age %>%
  mutate(rate = deaths / pop)
 
# Step 4: join weights + calculate weighted rates 
adj_rates <- data_age %>%
  left_join(std_pop, by="age_group") %>%
  mutate(weighted_rate = rate * weight)

#sum weighted rates --> age-adjusted rates
table1_adj <- adj_rates %>%
  group_by(race_eth, cod) %>%
  summarise(
    age_adj_rate = sum(weighted_rate) * 100000,
    .groups="drop"
  )

# Step 5: adding poisson CI 
table1_adj <- adj_rates %>%
  group_by(race_eth, cod) %>%
  summarise(
    adj_rate = sum(weighted_rate),
    deaths_total = sum(deaths),
    .groups="drop"
  ) %>%
  mutate(
    rate_100k = adj_rate * 100000,

    se_log = 1 / sqrt(deaths_total),

    lcl = rate_100k * exp(-1.96 * se_log),
    ucl = rate_100k * exp( 1.96 * se_log)
  )

#formatting
table1_adj <- table1_adj %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_100k, lcl, ucl
    )
  )
cod_levels <- c(
  "homicide",
  "mvc",
  "suicide",
  "other_injury",
  "other_disease",
  "all_cause"
)

cod_labels <- c(
  homicide = "Homicide",
  mvc = "Motor Vehicle Crashes",
  suicide = "Suicide",
  other_injury = "Other Injuries",
  other_disease = "Other Disease",
  all_cause = "Total (All Causes)"
)

table1_final <- table1_adj %>%
  mutate(
    cod = factor(cod, levels = cod_levels, ordered = TRUE),
    cod = recode(cod, !!!cod_labels)
  ) %>%
  arrange(cod)
library(tidyr)
library(dplyr)

table1_wide <- table1_final %>%
  select(cod, race_eth, rate_ci) %>%
  pivot_wider(
    names_from = race_eth,
    values_from = rate_ci
  ) %>%
  arrange(cod)

knitr::kable(
  table1_wide,
  caption = "Table 1. Annual mortality rates (95% CI) per 100,000 by cause of death and race/ethnic group. California males age 15–24 years, 2007–2020."
)


``` 

## Figures 

We are going to use some of the functions defined in our functions.R script to conduct our exploration.


```{r figures}

source("../Scripts/functions.R")

#-----------------------------------
# MERGE ONCE
#-----------------------------------

cdc <- cdc_mort %>%
  left_join(dof_data_grouped,
            by=c("year","sex","race_eth","age_group"))



yrs_0720 <- 2007:2020



#================================================
# FIGURE 1 — Male all-cause RR
#================================================
fig1 <- cdc %>%
  calc_rates(yrs_0720,"male","all_cause") %>%
  calc_rr()

plot_rr(fig1,"Male All-Cause Mortality Rate Ratios")



#================================================
# FIGURE 2 — Female all-cause RR
#================================================
fig2 <- cdc %>%
  calc_rates(yrs_0720,"female","all_cause") %>%
  calc_rr()

plot_rr(fig2,"Female All-Cause Mortality Rate Ratios")



#================================================
# FIGURE 3 — Male homicide rates
#================================================
fig3 <- cdc %>%
  calc_rates(yrs_0720,"male","homicide") %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_rate(fig3,"Male Homicide Mortality Rates")



#================================================
# FIGURE 4 — Male all-cause minus homicide RR
#================================================
fig4 <- cdc %>%
  filter(year %in% yrs_0720, sex=="male") %>%
  group_by(race_eth,age_group,cod) %>%
summarise(deaths=sum(deaths), .groups="drop") %>%
left_join(
  cdc %>%
    filter(year %in% yrs_0720, sex=="male") %>%
    group_by(race_eth,age_group) %>%
    summarise(pop=sum(population), .groups="drop"),
  by=c("race_eth","age_group")
) %>%
  pivot_wider(names_from=cod, values_from=deaths, values_fill=0) %>%
  mutate(deaths=all_cause-homicide,
         rate=deaths/pop*100000) %>%
  set_age_order() %>%
  calc_rr()

plot_rr(fig4,"Male Non-Homicide Mortality RR")



#================================================
# FIGURE 5 — Homicide yearly (15–24 males)
#================================================
fig5 <- cdc %>%
  filter(sex=="male",
         age_group %in% c("15-19","20-24"),
         cod=="homicide",
         year %in% 2007:2020) %>%
  group_by(year,race_eth) %>%
  summarise(deaths=sum(deaths),pop=sum(population),.groups="drop") %>%
  mutate(rate=deaths/pop*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_yearly(fig5,"Homicide Mortality Males 15–24")



#================================================
# FIGURE 6 — MVC rates by age
#================================================
fig6 <- cdc %>%
  calc_rates(yrs_0720,"male","mvc") %>%
  rowwise() %>%
  mutate(ci_log(rate,deaths)) %>%
  ungroup()

plot_rate(fig6,"Male Motor Vehicle Crash Mortality Rates")



#================================================
# FIGURE 7 — MVC yearly (15–24)
#================================================
fig7 <- cdc %>%
  filter(sex=="male",
         age_group %in% c("15-19","20-24"),
         cod=="mvc",
         year %in% 2007:2020) %>%
  group_by(year,race_eth) %>%
  summarise(deaths=sum(deaths),pop=sum(population),.groups="drop") %>%
  mutate(rate=deaths/pop*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_yearly(fig7,"MVC Mortality Males 15–24")

```



## Overlaid figure 

```{r fig1}
#Loading data 

cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 

age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84",
  "85-89", "90+"
)

dof_data_grouped <- dof_data_grouped %>% 
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  ) 

#Step 1 
figure1_data <- cdc_mort %>% 
  filter(
    year %in% 1999:2006,
    sex == "male", 
    cod == "all_cause"
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarise(
    death_count = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )


#Step 2
figure1_pop <- dof_data_grouped %>% 
  filter(
    year %in% 1999:2006,
    sex == "male"
  ) %>% 
  group_by(race_eth, age_group) %>%
  summarise(
    pop_1999_2006 = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) 


# Step 3:

figure1_rates <- figure1_data %>%
  left_join(
    figure1_pop,
    by = c("race_eth", "age_group")
  ) %>%
  mutate(
    rate = (death_count / pop_1999_2006) * 100000
  )

# Calculate rate ratios using NH White as reference within age group
rr_data <- figure1_rates %>%
  group_by(age_group) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = death_count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # exclude reference group from plotting
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1 / death_count + 1 / ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1 / death_count + 1 / ref_count))
  )


rr_data <- rr_data %>%
  mutate(
    age_group = factor(age_group, levels = age_levels, ordered = TRUE)
  )

ggplot(fig1,
       aes(x = age_group, y = rr, color = race_eth, group = race_eth)) +

  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +

  # MAIN PERIOD (solid)
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = rr_lcl, ymax = rr_ucl), width = 0.15) +

  # OVERLAY PERIOD (dashed)
  geom_line(
    data = rr_data,
    linetype = "dashed",
    size = 1
  ) +
  geom_point(
    data = rr_data,
    shape = 1,
    size = 2
  ) +
  geom_errorbar(
    data = rr_data,
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15,
    linetype = "dashed"
  ) +

  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Males",
    subtitle = "Solid = 2006–2020, Dashed = 1999–2006 | Reference: White",
    x = "Age (years)",
    y = "Rate Ratio"
  ) +
  theme_minimal()


```


