---
title: "Exploration"
output:
  html_document:
output_dir: ../Output
---
```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)
library(tidyr)

# Global options
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

```
# Exploring population data 

We are going to explore the distribution of our population data by individual ages and age categories, respectively. 

```{r explore_population, include = FALSE}
dof_data_grouped <- read.csv("../CleanData/dof_data_grouped.csv")
dof_data <- read.csv("../CleanData/dof_data.csv")
```

Let's examine the trends for the whole study period (data for 1999-2020):

```{r explore_age_groups, echo = FALSE} 
age_levels <- c(
  "0-4", "5-9", "10-14", "15-19", "20-24",
  "25-29", "30-34", "35-39", "40-44",
  "45-49", "50-54", "55-59", "60-64",
  "65-69", "70-74", "75-79", "80-84", "85-89", "90+"
)
agg_data <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  group_by(race_eth, age_group) %>%
  summarise(total_population = sum(population, na.rm = TRUE), .groups = "drop") %>%
  
  # convert to per 100,000
  mutate(pop_per_100k = total_population / 100000)

ggplot(agg_data, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()


``` 

Now let's look through the decades: 

```{r, echo = FALSE}
agg_period <- dof_data_grouped %>%
  mutate(age_group = factor(age_group, levels = age_levels, ordered = TRUE)) %>%
  mutate(period = case_when(
    year >= 2000 & year <= 2009 ~ "2000-2009",
    year >= 2010 & year <= 2020 ~ "2010-2020",
    TRUE ~ NA_character_
  )) %>%
  filter(!is.na(period)) %>%
  group_by(race_eth, age_group, period) %>%
  summarise(total_population = sum(population, na.rm = TRUE)) %>%
  ungroup() %>% 
   mutate(pop_per_100k = total_population / 100000)

ggplot(agg_period, aes(x = age_group, y = pop_per_100k, color = race_eth, group = race_eth)) +
  facet_wrap(~period, ncol = 1) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(
    title = "Overall Population by Age Group and Race/Ethnicity (per 100,000)",
    x = "Age Group",
    y = "Population per 100,000",
    color = "Race/Ethnicity"
  ) +
  theme_minimal()

```
# Sex differences 

```{r, include = FALSE}
population_table_wide <- dof_data_grouped %>%
  tidyr::pivot_wider(
    names_from = sex,
    values_from = population
  ) %>%
  arrange(age_group, year)

population_table_wide %>%
  kable(
    caption = "Population by Sex, Age Group, and Year"
  )

```


#All-cause mortality

To start, let's just look at all-cause mortality for males by racial/ethnic group. We are going to create a new dataframe for just this information: 
```{r all_cause_mort, echo = FALSE}
all_cause_mort <- read.csv("../CleanData/all_cause_mort.csv")

race_eth_all_cause <- all_cause_mort %>% 
  filter(Strata_Name %in% c("Female * Black", 
                            "Female * Hispanic", "Female * White",
                            "Male * Black", "Male * Hispanic",
                            "Male * White")) #excluding deaths for Multi-Race and Other/Unknown 

race_eth_all_cause <- race_eth_all_cause %>% #tidying
  separate(
    Strata_Name,
    into = c("sex", "race_eth"),
    sep = " \\* "
  ) %>% 
  select(-Strata) %>% 
  rename_with(tolower) %>% 
  mutate(sex = str_to_lower(sex),
         race_eth = str_to_lower(race_eth))
head(race_eth_all_cause)

```
Now that all of the naming conventions are the same as the population data, we can bring in the pop dataset and merge it with our race_eth_all_cause dataframe and calculate the death rate. We also need to apply the log-normal (Posson) CI for the rates.

```{r all_cause_with_pop, echo = FALSE}
dof_no_age <- read_csv("../CleanData/dof_no_age.csv")

all_cause_with_pop <- left_join(race_eth_all_cause,dof_no_age,
                                by = c("year", "sex", "race_eth"))

all_cause_with_pop <- all_cause_with_pop %>% 
  mutate( #mortality rate
          rate = count/population,
         rate_per_100k = rate*100000,
         
         # log-normal Poisson CI
    rate_lcl = rate * exp(-1.96 / sqrt(count)),
    rate_ucl = rate * exp( 1.96 / sqrt(count)),

    rate_lcl_100k = rate_lcl * 100000,
    rate_ucl_100k = rate_ucl * 100000)
head(all_cause_with_pop)
``` 


Let's summarize these trends: 

```{r all_c_no_age_trends, include = FALSE}
all_cause_table <- all_cause_with_pop %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_per_100k,
      rate_lcl_100k,
      rate_ucl_100k
    )
  )

#females
female_table <- all_cause_table %>%
  filter(sex == "female") %>%
  select(year, race_eth, rate_ci)

female_table %>%
  kable(
    caption = "ALl-Case Female Death Rate"
  )

#males 
male_table <- all_cause_table %>%
  filter(sex == "male") %>%
  select(year, race_eth, rate_ci)

male_table%>%
  kable(
    caption = "ALl-Case Male Death Rate"
  )
``` 

```{r echo = FALSE}
#sex stratified table
sex_stratified_table <- all_cause_table %>%
  select(year, sex, race_eth, rate_ci) %>%
  pivot_wider(
    names_from = sex,
    values_from = rate_ci
  )

sex_stratified_table%>%
  kable(
    caption = "ALl-Case Death Rate Stratified by Sex"
  )
```

*Mortality rates are expressed per 100,000 persons. 95% confidence intervals were calculated using the log-normal approximation to the Poisson distribution.*

Now, we are going to plot. We are going to use White as the referent category. 

```{r plot_no_age_rr, echo = FALSE}
rr_data <- all_cause_with_pop %>%
  group_by(year, sex) %>%
  mutate(
    ref_rate  = rate[race_eth == "white"],
    ref_count = count[race_eth == "white"]
  ) %>%
  ungroup() %>%
  filter(race_eth != "white") %>%   # drop reference from RR plot
  mutate(
    rr = rate / ref_rate,
    rr_lcl = rr * exp(-1.96 * sqrt(1/count + 1/ref_count)),
    rr_ucl = rr * exp( 1.96 * sqrt(1/count + 1/ref_count))
  )
rr_data
#females
ggplot(
  rr_data %>% filter(sex == "female"),
  aes(x = year, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Females",
    subtitle = "Reference group: White",
    x = "Year",
    y = "Rate Ratio"
  ) +
  theme_minimal()

#males 
ggplot(
  rr_data %>% filter(sex == "male"),
  aes(x = year, y = rr, color = race_eth, group = race_eth)
) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray50") +
  geom_line(size = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rr_lcl, ymax = rr_ucl),
    width = 0.15
  ) +
  labs(
    title = "Mortality Rate Ratios by Race/Ethnicity — Males",
    subtitle = "Reference group: White",
    x = "Year",
    y = "Rate Ratio"
  ) +
  theme_minimal()

```

# CDC Mortality Data 

```{r cdc_merge, include = FALSE}
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
head(cdc_mort)

#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")
head(dof_data)

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data,
                                by = c("year", "sex", "race_eth", "age_group"))

head(cdc_with_pop)
```


# Replicating tables and figures 
## 2006-2023 data 

To start, I am going to exclude the data from 2023, as we will need to figure out whether we are going to make two separate tables (standardizing to 2010 and 2020 population, with understanding that years after 2020 come from population *projections*, not estimates) or just include 2006-2020 data.

## Table 1

To replicate the table, we need to:  

1. subset deaths to males 15-24, years 2006-2020

2. create a population weighting variable for 2010

3. compute age-specific weights 

4. Join weights to calculate age-adjusted rates 

5. Calculate 95% CIs using log normal approximation to the Poisson distribution 


```{r}

#Step 1: keep deaths and population by age group 
data_age <- cdc_mort %>%
  filter(
    year %in% 2006:2020,
    age_group %in% c("15-19","20-24"),
    sex=="male"
  ) %>%
  group_by(race_eth, cod, age_group) %>%
  summarise(deaths = sum(deaths), .groups="drop") %>%
  left_join(
    dof_data_grouped %>%
      filter(
        year %in% 2006:2020,
        age_group %in% c("15-19","20-24"),
        sex=="male"
      ) %>%
      group_by(race_eth, age_group) %>%
      summarise(pop = sum(population), .groups="drop"),
    by=c("race_eth","age_group")
  )

#Step 2: get standard population weights 
std_pop <- dof_data_grouped %>%
  filter(
    year == 2010,
    age_group %in% c("15-19","20-24"),
    sex == "male"
  ) %>%
  group_by(age_group) %>%
  summarise(std_pop = sum(population), .groups="drop") %>%
  mutate(weight = std_pop / sum(std_pop))

# Step 3: compute age-specific rates
data_age <- data_age %>%
  mutate(rate = deaths / pop)
 
# Step 4: join weights + calculate weighted rates 
adj_rates <- data_age %>%
  left_join(std_pop, by="age_group") %>%
  mutate(weighted_rate = rate * weight)

#sum weighted rates --> age-adjusted rates
table1_adj <- adj_rates %>%
  group_by(race_eth, cod) %>%
  summarise(
    age_adj_rate = sum(weighted_rate) * 100000,
    .groups="drop"
  )

# Step 5: adding poisson CI 
table1_adj <- adj_rates %>%
  group_by(race_eth, cod) %>%
  summarise(
    adj_rate = sum(weighted_rate),
    deaths_total = sum(deaths),
    .groups="drop"
  ) %>%
  mutate(
    rate_100k = adj_rate * 100000,

    se_log = 1 / sqrt(deaths_total),

    lcl = rate_100k * exp(-1.96 * se_log),
    ucl = rate_100k * exp( 1.96 * se_log)
  )

#formatting
table1_adj <- table1_adj %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_100k, lcl, ucl
    )
  )
cod_levels <- c(
  "homicide",
  "mvc",
  "suicide",
  "other_injury",
  "other_disease",
  "all_cause"
)

cod_labels <- c(
  homicide = "Homicide",
  mvc = "Motor Vehicle Crashes",
  suicide = "Suicide",
  other_injury = "Other Injuries",
  other_disease = "Other Disease",
  all_cause = "Total (All Causes)"
)

table1_final <- table1_adj %>%
  mutate(
    cod = factor(cod, levels = cod_levels, ordered = TRUE),
    cod = recode(cod, !!!cod_labels)
  ) %>%
  arrange(cod)
library(tidyr)
library(dplyr)

table1_wide <- table1_final %>%
  select(cod, race_eth, rate_ci) %>%
  pivot_wider(
    names_from = race_eth,
    values_from = rate_ci
  ) %>%
  arrange(cod)

knitr::kable(
  table1_wide,
  caption = "Table 1. Annual mortality rates (95% CI) per 100,000 by cause of death and race/ethnic group. California males age 15–24 years, 2006–2020."
)


``` 
We are going to use some of the functions defined in our functions.R script to conduct our exploration.

```{r figures}

source("../Scripts/functions.R")

#-----------------------------------
# MERGE ONCE
#-----------------------------------

cdc <- cdc_mort %>%
  left_join(dof_data_grouped,
            by=c("year","sex","race_eth","age_group"))



yrs_0620 <- 2006:2020



#================================================
# FIGURE 1 — Male all-cause RR
#================================================
fig1 <- cdc %>%
  calc_rates(yrs_0620,"male","all_cause") %>%
  calc_rr()

plot_rr(fig1,"Male All-Cause Mortality Rate Ratios")



#================================================
# FIGURE 2 — Female all-cause RR
#================================================
fig2 <- cdc %>%
  calc_rates(yrs_0620,"female","all_cause") %>%
  calc_rr()

plot_rr(fig2,"Female All-Cause Mortality Rate Ratios")



#================================================
# FIGURE 3 — Male homicide rates
#================================================
fig3 <- cdc %>%
  calc_rates(yrs_0620,"male","homicide") %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_rate(fig3,"Male Homicide Mortality Rates")



#================================================
# FIGURE 4 — Male all-cause minus homicide RR
#================================================
fig4 <- cdc %>%
  filter(year %in% yrs_0620, sex=="male") %>%
  group_by(race_eth,age_group,cod) %>%
summarise(deaths=sum(deaths), .groups="drop") %>%
left_join(
  cdc %>%
    filter(year %in% yrs_0620, sex=="male") %>%
    group_by(race_eth,age_group) %>%
    summarise(pop=sum(population), .groups="drop"),
  by=c("race_eth","age_group")
) %>%
  pivot_wider(names_from=cod, values_from=deaths, values_fill=0) %>%
  mutate(deaths=all_cause-homicide,
         rate=deaths/pop*100000) %>%
  set_age_order() %>%
  calc_rr()

plot_rr(fig4,"Male Non-Homicide Mortality RR")



#================================================
# FIGURE 5 — Homicide yearly (15–24 males)
#================================================
fig5 <- cdc %>%
  filter(sex=="male",
         age_group %in% c("15-19","20-24"),
         cod=="homicide",
         year %in% 2006:2020) %>%
  group_by(year,race_eth) %>%
  summarise(deaths=sum(deaths),pop=sum(population),.groups="drop") %>%
  mutate(rate=deaths/pop*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_yearly(fig5,"Homicide Mortality Males 15–24")



#================================================
# FIGURE 6 — MVC rates by age
#================================================
fig6 <- cdc %>%
  calc_rates(yrs_0620,"male","mvc") %>%
  rowwise() %>%
  mutate(ci_log(rate,deaths)) %>%
  ungroup()

plot_rate(fig6,"Male Motor Vehicle Crash Mortality Rates")



#================================================
# FIGURE 7 — MVC yearly (15–24)
#================================================
fig7 <- cdc %>%
  filter(sex=="male",
         age_group %in% c("15-19","20-24"),
         cod=="mvc",
         year %in% 2006:2020) %>%
  group_by(year,race_eth) %>%
  summarise(deaths=sum(deaths),pop=sum(population),.groups="drop") %>%
  mutate(rate=deaths/pop*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,pop)) %>%
  ungroup()

plot_yearly(fig7,"MVC Mortality Males 15–24")

```


