---
title: "Replication"
author: "Ainsley Lotito"
date: "2026-02-06"
output:
  html_document:
output_dir: ../Output
---

```{r setup, include = FALSE}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(ggthemes)
library(knitr)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width=10, fig.height=5)

#Loading data 
dof_data_grouped <- read.csv("../CleanData/dof_data_grouped.csv")
dof_data <- read.csv("../CleanData/dof_data.csv")
all_cause_mort <- read.csv("../CleanData/all_cause_mort.csv")
dof_no_age <- read_csv("../CleanData/dof_no_age.csv")
cdc_mort <- read.csv("../CleanData/cdc_mortality.csv") 
head(cdc_mort)

#merging
cdc_with_pop <- left_join(cdc_mort,dof_data_grouped,
                                by = c("year", "sex", "race_eth", "age_group"))

cdc_with_pop <- cdc_with_pop %>% 
  mutate( #mortality rate
          rate = deaths/population,
         rate_per_100k = rate*100000,
         
         # log-normal Poisson CI
    rate_lcl = rate * exp(-1.96 / sqrt(deaths)),
    rate_ucl = rate * exp( 1.96 / sqrt(deaths)),

    rate_lcl_100k = rate_lcl * 100000,
    rate_ucl_100k = rate_ucl * 100000) 

```

All cause-mortality 
```{r}
race_eth_all_cause <- all_cause_mort %>% 
  filter(Strata_Name %in% c("Female * Black", 
                            "Female * Hispanic", "Female * White",
                            "Male * Black", "Male * Hispanic",
                            "Male * White")) #excluding deaths for Multi-Race and Other/Unknown 

race_eth_all_cause <- race_eth_all_cause %>% #tidying
  separate(
    Strata_Name,
    into = c("sex", "race_eth"),
    sep = " \\* "
  ) %>% 
  select(-Strata) %>% 
  rename_with(tolower) %>% 
  mutate(sex = str_to_lower(sex),
         race_eth = str_to_lower(race_eth))

all_cause_with_pop <- left_join(race_eth_all_cause,dof_no_age,
                                by = c("year", "sex", "race_eth"))

all_cause_with_pop <- all_cause_with_pop %>% 
  mutate( #mortality rate
          rate = count/population,
         rate_per_100k = rate*100000,
         
         # log-normal Poisson CI
    rate_lcl = rate * exp(-1.96 / sqrt(count)),
    rate_ucl = rate * exp( 1.96 / sqrt(count)),

    rate_lcl_100k = rate_lcl * 100000,
    rate_ucl_100k = rate_ucl * 100000)


```


# Replication of tables/figures from Vaca et al. (2011)

## Table 1

I think the easiest thing to match up is the table of annual mortality rates per 100,000 by cause of death and race/ethnic group among California males age 15-24 years, 1999-2006. 

To replicate the table, we need to:  

1. subset to males 15-24

2. collapse years 1999-2006 

3. use population data from 1999-2006 to calculate overall rates


```{r table_1_replication, echo = FALSE} 
table1_2000_pop <- dof_data_grouped %>% 
  filter(
    year %in% 1999:2006,
    age_group %in% c("15-19", "20-24"),
    sex == "male"
  ) %>% 
  group_by(race_eth) %>%
  summarise(
    pop_1999_2006 = sum(population, na.rm = TRUE),
    .groups = "drop"
  )



#selecting just the deaths for each age group from CDC data 
table1_data <- cdc_mort %>% 
  filter(
    year %in% 1999:2006, 
    age_group %in% c("15-19", "20-24"),
    sex == "male"
  ) %>% 
  group_by(race_eth, cod) %>% 
  summarise(
    death_count = sum(deaths, na.rm = TRUE),
    .groups = "drop"
  )
table1_data

#merging population data 
table1_data <- left_join(table1_data,
                         table1_2000_pop,
                         by = "race_eth")

table1_data



table1_data <- table1_data %>%
  mutate(
    rate_per_100k = death_count / pop_1999_2006 * 100000,
    rate_lcl = (qchisq(0.025, 2 * death_count) / 2) / pop_1999_2006 * 100000,
    rate_ucl = (qchisq(0.975, 2 * (death_count + 1)) / 2) / pop_1999_2006 * 100000
  )


table1_data %>% 
  group_by(cod) %>% 
  select(cod, race_eth,rate_per_100k)

table1_long <- table1_data %>%
  mutate(
    cod = factor(
      cod,
      levels = c(
        "homicide",
        "mvc",
        "suicide",
        "other_injury",
        "other_disease",
        "all_cause"
      )
    ),
    race_eth = factor(
      race_eth,
      levels = c("hispanic", "white", "black")
    ),
    
   # rate per 100k
    rate_100k = death_count / pop_1999_2006 * 100000,

    # Poisson-based CI
    lcl_100k = (qchisq(0.025, 2 * death_count) / 2) / pop_1999_2006 * 100000,
    ucl_100k = (qchisq(0.975, 2 * (death_count + 1)) / 2) / pop_1999_2006 * 100000
  )

#format rate + CI 
table1_long <- table1_long %>%
  mutate(
    rate_ci = sprintf(
      "%.1f (%.1f–%.1f)",
      rate_100k, lcl_100k, ucl_100k
    )
  )


table1_final <- table1_long %>%
  select(cod, race_eth, rate_ci) %>%
  pivot_wider(
    names_from = race_eth,
    values_from = rate_ci
  ) %>%
  arrange(cod)



table1_final

#poisson-based CI: 
#summary_tbl <- summary_tbl %>%
 # mutate(
 #   rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population / years,
   # rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population / years,
   # rate_lcl_100k = rate_lcl * 100000,
  #  rate_ucl_100k = rate_ucl * 100000
  #)


```

## Figure 1 

Next, we are going to rep 


## Figure 5 

```{r figure_5, echo = FALSE, fig.width=9, fig.height= 4}
fig5_data <- cdc_with_pop %>%
  filter(
    sex == "male",
    age_group %in% c("15-19", "20-24"),
    cod == "homicide",
    year <= 2006
  )

fig5_yearly <- fig5_data %>%
  group_by(year, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )

fig5_yearly <- fig5_yearly %>%
  mutate(
    rate = deaths / population,
    rate_per_100k = rate * 100000
  )

fig5_yearly <- fig5_yearly %>%
  mutate(
    rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population * 100000,
    rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population * 100000
  )

fig5_yearly %>% arrange(desc(rate_per_100k))



ggplot(fig5_yearly,
       aes(x = year, y = rate_per_100k, color = race_eth)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_errorbar(
    aes(ymin = rate_lcl, ymax = rate_ucl),
    width = 0.2
  ) +
  scale_x_continuous(breaks = 1999:2006) +
  scale_y_continuous(
    name = "Mortality rate per 100,000"
  ) +
  scale_color_manual(
    values = c(
      "hispanic" = "#E69F00",
      "black" = "#000000",
      "white" = "#56B4E9"
    ),
    labels = c(
      "hispanic" = "Latino",
      "black" = "Non-Latino Black",
      "white" = "Non-Latino White"
    )
  ) +
  labs(
    title= "Homicide Mortality Rate, Males 15-24 y/o, 2000-2006",
    x = "Year",
    color = "Race/Ethnicity"
  ) +
 theme_minimal()

```

That's still not great, but let's see about replicating figure 7?? 

```{r figure_7, echo = FALSE, fig.width=9, fig.height= 4}
fig7_data <- cdc_with_pop %>%
  filter(
    sex == "male",
    age_group %in% c("15-19", "20-24"),
    cod == "mvc",
    year <= 2006
  )

fig7_yearly <- fig7_data %>%
  group_by(year, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )

fig7_yearly <- fig7_yearly %>%
  mutate(
    rate = deaths / population,
    rate_per_100k = rate * 100000
  )

fig7_yearly <- fig7_yearly %>%
  mutate(
    rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population * 100000,
    rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population * 100000
  )

fig7_yearly %>% arrange(desc(rate_per_100k))




pd <- position_dodge(width = 0.3)

ggplot(fig7_yearly, aes(x = year, y = rate_per_100k, color = race_eth, linetype = race_eth)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, position = pd) +
  geom_errorbar(aes(ymin = rate_lcl, ymax = rate_ucl), width = 0.2, position = pd) +
  scale_x_continuous(breaks = 1999:2006) +
  scale_y_continuous(name = "Mortality rate per 100,000") +
  
  # Custom colors
  scale_color_manual(
    values = c(
      "hispanic" = "black",       # Latino = solid black
      "black" = "#000000",        # Non-Latino Black = dotted
      "white" = "black"         # Non-Latino White = dashed blue
    ),
    labels = c(
      "hispanic" = "Latino",
      "black" = "Non-Latino Black",
      "white" = "Non-Latino White"
    )
  ) +
  
  # Custom linetypes
  scale_linetype_manual(
    values = c(
      "hispanic" = "solid",       # solid line
      "black" = "dotted",         # dotted (smaller dashes)
      "white" = "dashed"          # dashed
    )
  ) +
  
  labs(
    title = "Motor vehicle crash mortality rates males 15-24",
    x = "Year",
    color = "Race/Ethnicity",
    linetype = "Race/Ethnicity"
  ) +
  theme_minimal()

```



We are going to do the same thing, except this time we will extend the data to go through years 2006-2021:

```{r figure_7_extended, echo = FALSE, fig.width=11, fig.height= 6}
fig7x_data <- cdc_with_pop %>%
  filter(
    sex == "male",
    age_group %in% c("15-19", "20-24"),
    cod == "mvc",
    year >= 2006, 
    year < 2021
  )

fig7x_yearly <- fig7x_data %>%
  group_by(year, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = TRUE),
    population = sum(population, na.rm = TRUE),
    .groups = "drop"
  )

fig7x_yearly <- fig7x_yearly %>%
  mutate(
    rate = deaths / population,
    rate_per_100k = rate * 100000
  )

fig7x_yearly <- fig7x_yearly %>%
  mutate(
    rate_lcl = (qchisq(0.025, 2 * deaths) / 2) / population * 100000,
    rate_ucl = (qchisq(0.975, 2 * (deaths + 1)) / 2) / population * 100000
  )

fig7x_yearly %>% arrange(desc(rate_per_100k))




pd <- position_dodge(width = 0.3)

ggplot(fig7x_yearly, aes(x = year, y = rate_per_100k, color = race_eth, linetype = race_eth)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, position = pd) +
  geom_errorbar(aes(ymin = rate_lcl, ymax = rate_ucl), width = 0.2, position = pd) +
  scale_x_continuous(breaks = 2006:2023) +
  scale_y_continuous(name = "Mortality rate per 100,000",
                     limits = c(0, 60) ) +
  
  # Custom colors
  scale_color_manual(
    values = c(
      "hispanic" = "black",       # Latino = solid black
      "black" = "#000000",        # Non-Latino Black = dotted
      "white" = "black"         # Non-Latino White = dashed blue
    ),
    labels = c(
      "hispanic" = "Latino",
      "black" = "Non-Latino Black",
      "white" = "Non-Latino White"
    )
  ) +
  
  # Custom linetypes
  scale_linetype_manual(
    values = c(
      "hispanic" = "solid",       # solid line
      "black" = "dotted",         # dotted (smaller dashes)
      "white" = "dashed"          # dashed
    )
  ) +
  
  labs(
    title = "Motor vehicle crash mortality rates males 15-24",
    x = "Year",
    color = "Race/Ethnicity",
    linetype = "Race/Ethnicity"
  ) +
  theme_minimal()

fig7x_table <- fig7x_yearly %>%
  select(year, race_eth, rate_per_100k) %>%
  pivot_wider(
    names_from = race_eth,       # new columns from race_eth
    values_from = rate_per_100k  # values to fill in
  )

fig7x_table
```
