---
title: "Underlying Exploration"
output_dir: Output 
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
--- 
```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

source("../Scripts/functions.R")
```
 
 
 
This document contains the analyses used to replicate Vaca et al. (2011). Datasets are pulled from the `/Cleandata` directory, which were produced in the `/scripts/Cleaning.Rmd` document.  


# All-Cause Mortality  

We are going to start by importing our all-cause data and plot the rates by year.  
```{r}
all_cause <- read.csv("../CleanData/all_cause_UCD.csv")
head(all_cause)

#adding pop data 
dof_data<- read_csv("../CleanData/dof_data_grouped.csv")

#merging
all_cause <- left_join(all_cause,dof_data,
                                by = c("year", "sex", "race_eth", "age_group")) 
head(all_cause) 

#making a rate per 100k variable 

all_cause <- all_cause %>% 
  mutate(rate = deaths/(population*100000)) 

#plotting by year 
ggplot(all_cause, aes(year, rate, col= race_eth))+
  facet_wrap(~age_group)+
  geom_line()  

#table of all ages by year, sex, and race/eth
all_cause %>% 
    group_by(race_eth, year, sex) %>%
    summarise(
      deaths = sum(deaths, na.rm=TRUE),
      pop    = sum(population, na.rm=TRUE),
      .groups="drop"
    ) %>%
    mutate(rate = deaths/pop*100000) 

youths <- c("15-19", "20-24")

#table of 15-24 year old males by year 
all_cause %>% 
  filter(sex == "male", 
         age_group %in% youths ) %>% 
    group_by(race_eth, year, sex) %>%
    summarise(
      deaths = sum(deaths, na.rm=TRUE),
      pop    = sum(population, na.rm=TRUE),
      .groups="drop"
    ) %>%
    mutate(rate = deaths/pop*100000) %>% 
  select(-deaths, -pop)

```  

# Descriptive Analysis 

## Total Death Count 

How many deaths occurred over the study period?
```{r}  
all_cause %>% 
  filter(year %in% 2007:2020) %>% 
  summarize(total_deaths = sum(deaths, na.rm = TRUE))

``` 
## Age-adjusted mortality rates 


## Fig 1- Male 

Male all-cause mortality rate ratios comparing Hispanic and non-Hispanic black mortality to non-Hispanic white mortality, by 5 year age groups with 95% CI, California, 2007-2020.
```{r}
yrs_0720 <- 2007:2020
fig1 <- calc_rates(all_cause,yrs_0720,"male","all_cause") %>%
  calc_rr() 

plot_rr(fig1,NULL)  

#trying again, this time for the other years 
yrs_9906 <-c(1999:2006)
fig1_old <- calc_rates(all_cause,yrs_9906,"male","all_cause") %>%
  calc_rr() 

plot_rr(fig1_old,NULL)  

``` 


Looking through the numbers: 

```{r}
tables_by_race <- fig1 %>%
  mutate(rr_ci = sprintf("%.2f (%.2f, %.2f)", rr, rr_lcl, rr_ucl)) %>%
  select(race_eth, age_group, rr_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(tables_by_race) <- unique(fig1$race_eth)

tables_by_race

``` 


## Fig 2- Female

Female all cause mortality rate ratios comparing Hispanic and non-Hispanic black mortality to non-Hispanic white mortality, by 5 year age groups with 95% CI, California, 2007-2020.

```{r}
fig2 <- all_cause %>%
  calc_rates(yrs_0720,"female","all_cause") %>%
  calc_rr()

plot_rr(fig2, NULL)

fig2_plot <- plot_rr(fig2, NULL)
ggsave(
  filename = "../Output/fig2.png",
  plot = fig2_plot,
  width = 10,
  height = 6
)

``` 
There is a slight problem here, because we are missing the 5-9 counts for Blacks most of the years due to data suppression. Because of this, we will eventually need to exclude this group. 

Looking through the numbers: 

```{r}
tables_by_race <- fig2 %>%
  mutate(rr_ci = sprintf("%.2f (%.2f, %.2f)", rr, rr_lcl, rr_ucl)) %>%
  select(race_eth, age_group, rr_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(tables_by_race) <- unique(fig1$race_eth)

tables_by_race

```  


# Homicide 

We will pull two cleaned datasets for this, including the homicide counts by age group for the whole study period, and the homicide counts by year. 

## Figure 3 

```{r}
hom_age <- read.csv("../CleanData/homicide_UCD_age_groups.csv")
head(hom_age) 

fig3 <- hom_age %>%
    mutate(rate = deaths/population*100000) %>%
    set_age_order() %>% 
   rowwise() %>%
  mutate(ci_exact(deaths,population)) %>%
  ungroup()

plot_rate(fig3,NULL) 

fig3_plot <- plot_rate(fig3, NULL)
ggsave(
  filename = "../Output/fig3UCD.png",
  plot = fig3_plot,
  width = 10,
  height = 6
)
``` 

By number: 
```{r}
tables_by_race <- fig3 %>%
  mutate(r_ci = sprintf("%.2f (%.2f, %.2f)", rate, lcl, ucl)) %>%
  select(race_eth, age_group, r_ci) %>%
  arrange(race_eth, age_group) %>%
  group_split(race_eth)

names(tables_by_race) <- unique(fig3$race_eth)

tables_by_race 
``` 



## Figure 5: Male 15-24  

Homicide mortality rates, among males 15-24 years of age, comparing Hispanic, non-Hispanic black, and non-Hispanic white, by year with 95% CI, California 2007-2020.
```{r} 
fig5data <- read.csv("../CleanData/homicide_UCD_yearly.csv")
fig5 <- fig5data %>%
  mutate(rate=deaths/population*100000) %>%
  rowwise() %>%
  mutate(ci_exact(deaths,population)) %>%
  ungroup()

plot_yearly(fig5, NULL) 

fig5_plot <- plot_yearly(fig5, NULL) 
ggsave(
  filename = "../Output/fig5UCD.png",
  plot = fig5_plot,
  width = 10,
  height = 6
)

``` 
