---
title: "Cleaning"
output_dir: Output 
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    toc_float: true
--- 
```{r setup}
# Load packages
library(tidyverse)
library(ggplot2) 
library(readxl) 
library(lubridate)
library(janitor)

# Global options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

source("../Scripts/functions.R")
```
 
# All-Cause Mortality by Year 

```{r}
#H 
all_cause_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/All_Cause/H.xlsx",
                          col_names = TRUE)
all_cause_h <- H_ucd_clean(all_cause_h, "all_cause")  
glimpse(all_cause_h)

#NH 
all_cause_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/All_Cause/NH.xlsx",
                          col_names = TRUE) 

all_cause_nh <- NH_ucd_clean(all_cause_nh, "all_cause") 
glimpse(all_cause_nh)  

#Binding rows 
all_cleaned <- bind_rows(
  all_cause_h,
  all_cause_nh
)

glimpse(all_cleaned)
#Cleaning final merge

all_cleaned <- all_cleaned %>%
  filter(age_group != "Not Stated") %>% 
  mutate(
    sex = str_to_lower(sex),
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ), 
    age_group = str_trim(age_group),
    age_group = case_when(
      age_group %in% c("< 1 year", "1-4 years") ~ "0-4",
      age_group == "5-9 years"   ~ "5-9",
      age_group == "10-14 years" ~ "10-14",
      age_group == "15-19 years" ~ "15-19",
      age_group == "20-24 years" ~ "20-24",
      age_group == "25-29 years" ~ "25-29",
      age_group == "30-34 years" ~ "30-34",
      age_group == "35-39 years" ~ "35-39",
      age_group == "40-44 years" ~ "40-44",
      age_group == "45-49 years" ~ "45-49",
      age_group == "50-54 years" ~ "50-54",
      age_group == "55-59 years" ~ "55-59",
      age_group == "60-64 years" ~ "60-64",
      age_group == "65-69 years" ~ "65-69",
      age_group == "70-74 years" ~ "70-74",
      age_group == "75-79 years" ~ "75-79",
      age_group == "80-84 years" ~ "80-84",
      age_group == "85-89 years" ~ "85-89",
      age_group %in% c("90-94 years", "95-99 years", "100+ years") ~ "90+",
      TRUE ~ NA_character_
    )
  )
all_cleaned <- all_cleaned %>% 
  group_by(year, sex, age_group, cod, race_eth) %>%
  summarise(
    deaths = sum(deaths, na.rm = FALSE),
    .groups = "drop"
  ) %>% 
  arrange(year,age_group)

# checking NA values
all_cleaned %>% 
  filter(is.na(deaths))  

head(all_cleaned)

write.csv(all_cleaned, "../CleanData/all_cause_UCD.csv")

  
``` 

The only NA death values (they were suppressed due to data constraints) are among Black children, usually 5-9 years old. This makes some sense, as they are the smallest racial group in CA.  



# Homicide  

## By age
```{r}
# By Age Group 
hom_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Age/H.xlsx", col_names = TRUE)
hom_h <- ucd_age_clean_h(hom_h, "homicide")  
glimpse(hom_h) 

#we have 23 obs, which is equivalent to one hispanic row for each age group. looks good. 

#NH
hom_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Age/NH.xlsx", col_names = TRUE)
hom_nh <- ucd_age_clean_nh(hom_nh, "homicide")  
glimpse(hom_nh)

age_homicide <-  bind_rows(
  hom_nh,
  hom_h
) 


#Cleaning final merge

age_homicide <- race_age_format(age_homicide) 
glimpse(age_homicide)

#adding population data 
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarize(population = sum(population))
head(pop) 

age_hom2 <- age_homicide %>%
  left_join(pop, by = c("race_eth", "age_group"))
head(age_hom2)

write.csv(age_hom2, "../CleanData/homicide_UCD_age_groups.csv")
```

## By year: 
```{r}
#By Year (15-24 year olds)
hom_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Year/H.xlsx", col_names = TRUE)
hom_h <- ucd_year_clean_h(hom_h, "homicide")  
glimpse(hom_h) 

#NH
hom_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Homicide/Year/NH.xlsx", col_names = TRUE)
hom_nh <- ucd_year_clean_nh (hom_nh, "homicide")  
glimpse(hom_nh)

year_homicide <-  bind_rows(
  hom_nh,
  hom_h
) 


#Cleaning final merge

year_homicide <- year_homicide %>% 
  mutate(
    race_eth = case_when(
      race == "Black or African American" ~ "black",
      race == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    )
  )
glimpse(year_homicide) 
unique(year_homicide$race_eth)

#adding population data 
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    age_group %in% c("15-19","20-24"),
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, year) %>% 
  summarize(population = sum(population))
head(pop) 

year_hom2 <- year_homicide %>%
  left_join(pop, by = c("race_eth", "year"))
head(year_hom2)

write.csv(year_hom2, "../CleanData/homicide_UCD_yearly.csv")

```

# MVC 

## By Age 

```{r}
mvc_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Age/NH.xlsx", col_names = TRUE)
head(mvc_nh)

mvc_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Age/H.xlsx", col_names = TRUE)
head(mvc_h) 

mvc_nh <- mvc_nh %>% 
  filter(Sex == "Male") %>% 
  ucd_age_clean_nh( "mvc")

mvc_h <- mvc_h %>% 
  filter(Sex == "Male") %>%
  ucd_age_clean_h( "mvc") 

age_mvc <-  bind_rows(
  mvc_nh,
  mvc_h
) 
head(age_mvc)

age_mvc <- race_age_format(age_mvc)
age_mvc %>% 
  group_by(race_eth) %>% 
  slice(1:3)

#adding population data
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, age_group) %>% 
  summarize(population = sum(population))
head(pop) 

age_mvc <- age_mvc %>%
  left_join(pop, by = c("race_eth", "age_group"))
head(age_mvc)
write.csv(age_mvc, "../CleanData/mvc_UCD_age.csv")
``` 

## By Year 

```{r}
#NH
mvc_y_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Year/NH.xlsx", col_names = TRUE)
head(mvc_y_nh)
mvc_y_nh <- ucd_year_clean_nh(mvc_y_nh, "mvc") 

#H
mvc_y_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/MVC/Year/H.xlsx", col_names = TRUE)
head(mvc_y_h)
mvc_y_h <- ucd_year_clean_h(mvc_y_h, "mvc") 

#Binding
year_mvc <-  bind_rows(
  mvc_y_nh,
  mvc_y_h
) 


#Cleaning final merge

year_mvc <- year_mvc %>% 
  mutate(
    race_eth = case_when(
      race == "Black or African American" ~ "black",
      race == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    )
  )

glimpse(year_mvc) 
unique(year_mvc$race_eth) 

#adding population data 
#(for all of 2007-2020, by age group)
pop <- read.csv("../CleanData/dof_data_grouped.csv")
head(pop) 

pop <- pop %>% 
  filter(
    sex == "male",
    age_group %in% c("15-19","20-24"),
    year %in% c(2007:2020)
  ) %>% 
  group_by(race_eth, year) %>% 
  summarize(population = sum(population))
head(pop) 

year_mvc <- year_mvc %>%
  left_join(pop, by = c("race_eth", "year"))
head(year_mvc)

write.csv(year_mvc, "../CleanData/mvc_UCD_yearly.csv")


```

# COD Males 15-24 2007-2020 

For our table 1 data, we need to make a larger dataframe that combines all the causes of death by race/ethnicity among just males aged 15-24 over the study period.  

We will start by pulling up the dataframes that already meet these conventions. We are also going to just retain cod, race_eth, and deaths.

```{r}
#mvc
age_mvc <- age_mvc %>% 
  filter(age_group %in% c("15-19", "20-24")) %>% 
  select(cod, race_eth, age_group, deaths) 
head(age_mvc)
#homicide
age_hom <- age_hom2 %>% 
  filter(age_group %in% c("15-19", "20-24")) %>% 
  select(cod, race_eth, age_group, deaths)
head(age_hom)
#all cause
head(all_cleaned) 
all_cause_table1 <- all_cleaned %>% 
  filter(year %in% c(2007:2020),
         sex == "male", 
         age_group %in% c("15-19", "20-24")) %>% 
  group_by(age_group, race_eth, cod) %>% 
  summarize(deaths = sum(deaths))
head(all_cause_table1)
``` 

Next we have to clean disease and injury accordingly: 
```{r} 

#function for later (import labeled the years strangely): 

#Disease 

#NH
dis_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Disease/Age/NH.xlsx", col_names = TRUE)
head(dis_nh)
dis_nh <- ucd_age_clean_nh(dis_nh, "disease")  


#H
dis_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Disease/Age/H.xlsx", col_names = TRUE)
head(dis_h)
dis_h <- ucd_age_clean_h(dis_h, "disease") 
 sum(is.na(dis_nh$Deaths) | dis_nh$Deaths == 0, na.rm = FALSE)
  

#Binding
age_dis <-  bind_rows(
 dis_h,
 dis_nh
) 

#Cleaning final merge
age_dis <- age_dis %>% 
  mutate(
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ),
    age_group = gsub(" years", "", age_group),
      age_group = trimws(age_group)
  )

head(age_dis)

# --------------------- 
#Injury  
#NH
inj_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Injuries/Age/NH.xlsx", col_names = TRUE)
head(inj_nh)
inj_nh <- ucd_age_clean_nh(inj_nh, "injury") 

#H
inj_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Injuries/Age/H.xlsx", col_names = TRUE)
head(inj_h)
inj_h <- ucd_age_clean_h(inj_h, "injury") 
 sum(is.na(inj_nh$Deaths) | inj_nh$Deaths == 0, na.rm = FALSE)

#Binding
age_inj <-  bind_rows(
 inj_h,
 inj_nh
) 

#Cleaning final merge
age_inj <- age_inj %>% 
  mutate(
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ),
    age_group = gsub(" years", "", age_group),
      age_group = trimws(age_group)
  )
  

head(age_inj) 

# --------------------- 
#Suicide  
#NH
sui_nh <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Suicide/Age/NH.xlsx", col_names = TRUE)
head(sui_nh)
sui_nh  <- ucd_age_clean_nh(sui_nh , "suicide") 

#H
sui_h <- read_excel("../RawData/cdcWONDER/UnderlyingCOD99_20/Suicide/Age/H.xlsx", col_names = TRUE)
head(sui_h)
sui_h <- ucd_age_clean_h(sui_h, "suicide") 
 sum(is.na(sui_h$Deaths) | sui_h$Deaths == 0, na.rm = FALSE)

#Binding
age_sui <-  bind_rows(
 sui_h,
 sui_nh
) 

#Cleaning final merge
age_sui <- age_sui %>% 
 mutate(
    race_eth = case_when(
      race_eth == "Black or African American" ~ "black",
      race_eth == "White" ~ "white",
      race_eth == "Hispanic" ~ "hispanic",
      TRUE ~ race_eth
    ),
    age_group = gsub(" years", "", age_group),
      age_group = trimws(age_group)
  )
  

head(age_sui) 

``` 

Now that we have our datasets in the right configuration, we can bind all of them to make one large dataset with all causes of death among the population (15-24 year old males).

```{r}
table1 <- bind_rows(
  all_cause_table1,
  age_hom,
  age_mvc,
  age_inj,
  age_sui,
  age_dis
) 

head(table1) 
glimpse(table1) 

write_csv(table1, "../CleanData/table1.csv")
```